---
title: '20.4'
author: '520004915'
date: "2024-04-19"
output: html_document
---
#load packages
```{r setup, include=FALSE}
library(caret)
library(cvTools)
library(dplyr)
library(e1071)
library(GEOquery)
library(ggplot2)
library(glmnet)
library(limma)
library(pheatmap)
library(randomForest)
library(stringr)
library(tidyverse)
rm(list = ls())
```
#laod data sets & cleaning & gene probe sub
## 1. 'GSE36059'
```{r}
gse1 <- 'GSE36059'
eSet1 <- getGEO('GSE36059')
eSet1 <- eSet1$GSE36059_series_matrix.txt.gz
exp1 <- exprs(eSet1)
phe1 <- pData(phenoData(eSet1))
feature1 <- fData(eSet1)
```
### clean feature
```{r}
f1 <-feature1%>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
#f1[grepl('///', f1$`Gene Symbol`), ] after checking, i found they are different genes so i remove
f1$'Gene Symbol' <- sapply(strsplit(f1$'Gene Symbol', ' /// '), function(x) x[1])
```
### probe sub (add gene sybol)
```{r}
gene_symbols1 <- f1[, c(1, 11)]
expr1 <- as.data.frame(exp1)
expr1[, 'ID'] <- rownames(expr1)
expr1 <- merge(x = expr1, y = gene_symbols1, by.x = 'ID')
```
### remove duplication (keep mean)
```{r}
#table(duplicated(expr1$`Gene Symbol`)) #22902 suplication
expre1 <- limma::avereps(expr1[,-c(1,ncol(expr1))], ID = expr1$`Gene Symbol`)
expre1 <- as.data.frame(expre1)
expression1 <- expre1[, -c(1,ncol(expre1))]
dim(expression1)
```
### use expr to filter pheno
```{r}
pheno1 <- phe1[rownames(phe1) %in% colnames(expression1), ]
```
### set group
```{r}
pheno1 <- pheno1 %>% 
  filter(`diagnosis (tcmr,abmr,mixed,non-rejecting):ch1` %in% c('ABMR', 'TCMR'))%>%
  mutate(Outcome = `diagnosis (tcmr,abmr,mixed,non-rejecting):ch1`)
table(pheno1$Outcome)
```
## 2. 'GSE129166'
```{r}
gse2 <- 'GSE129166'
eSet2 <- getGEO('GSE129166')
eSet2 <- eSet2$GSE129166_series_matrix.txt.gz
exp2 <- exprs(eSet2)
phe2 <- pData(phenoData(eSet2))
feature2 <- fData(eSet2)
```
### clean feature
```{r}
f2 <-feature2%>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f2$'Gene Symbol' <- sapply(strsplit(f2$'Gene Symbol', ' /// '), function(x) x[1])
```
### probe sub (add gene sybol)
```{r}
gene_symbols2 <- f2[, c(1, 11)]
expr2 <- as.data.frame(exp2)
expr2[, 'ID'] <- rownames(expr2)
expr2 <- merge(x = expr2, y = gene_symbols2, by.x = 'ID')
```
### remove duplication (keep mean)
```{r}
#table(duplicated(expr2$`Gene Symbol`)) #22902 suplication
expre2 <- limma::avereps(expr2[,-c(1,ncol(expr2))], ID = expr2$`Gene Symbol`)
expre2 <- as.data.frame(expre2)
expression2<- expre2[, -c(1,ncol(expre2))]
dim(expression2)
```
### use expr to filter pheno
```{r}
pheno2 <- phe2[rownames(phe2) %in% colnames(expression2), ]
```
### set group
```{r}
pheno2 <- pheno2 %>%
  mutate(
    ABMR_Status = ifelse(`abmr (no:ch1` == '0_Yes:1): 1', 'ABMR', 'ABMR Stable'),
    TCMR_Status = ifelse(`tcmr (no:ch1` == '0_borderline:1_TCMR:2): 1', 'TCMR', 'TCMR Stable'),
    Outcome = ifelse(ABMR_Status == 'ABMR' & TCMR_Status == 'TCMR Stable', 'ABMR',
                     ifelse(ABMR_Status == 'ABMR Stable' & TCMR_Status == 'TCMR', 'TCMR', NA))
  )%>%
  na.omit(Outcome)
table(pheno2$Outcome)
```
## 3. GSE192444
```{r}
gse3 <- 'GSE192444'
eSet3 <- getGEO('GSE192444')
eSet3 <- eSet3$GSE192444_series_matrix.txt.gz
exp3 <- exprs(eSet3)
phe3<- pData(phenoData(eSet3))
feature3 <- fData(eSet3)
```
### clean feature
```{r}
f3 <-feature3%>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f3$'Gene Symbol' <- sapply(strsplit(f3$'Gene Symbol', ' /// '), function(x) x[1])
```
### probe sub (add gene sybol)
```{r}
gene_symbols3 <- f3[, c(1, 15)]
expr3 <- as.data.frame(exp3)
expr3[, 'ID'] <- rownames(expr3)
expr3 <- merge(x = expr3, y = gene_symbols3, by.x = 'ID')
```
### remove duplication (keep mean)
```{r}
table(duplicated(expr3$`Gene Symbol`)) #29822  suplication
expre3 <- limma::avereps(expr3[,-c(1,ncol(expr3))], ID = expr3$`Gene Symbol`)
expre3 <- as.data.frame(expre3)
expression3<- expre3[, -c(1,ncol(expre3))]
dim(expression3)
```
### use expr to filter pheno
```{r}
pheno3 <- phe3[rownames(phe3) %in% colnames(expression3), ]
```
### set group
```{r}
pheno3 <- pheno3 %>%
  mutate(
    Outcome = ifelse(`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1` == 'ABMR','ABMR',
                     ifelse(`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1`=='TCMR', 'TCMR', NA))
  )%>%
  na.omit(Outcome)
table(pheno3$Outcome)
```
## 4. GSE124203
```{r}
gse4 <- 'GSE124203'
eSet4 <- getGEO('GSE124203')
eSet4 <- eSet4$GSE124203_series_matrix.txt.gz
exp4 <- exprs(eSet4)
phe4<- pData(phenoData(eSet4))
feature4 <- fData(eSet4)
```
### clean feature
```{r}
f4 <-feature4%>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f4$'Gene Symbol' <- sapply(strsplit(f4$'Gene Symbol', ' /// '), function(x) x[1])
```
### probe sub (add gene sybol)
```{r}
gene_symbols4 <- f4[, c(1, 15)]
expr4 <- as.data.frame(exp4)
expr4[, 'ID'] <- rownames(expr4)
expr4 <- merge(x = expr4, y = gene_symbols4, by.x = 'ID')
```
### remove duplication (keep mean)
```{r}
table(duplicated(expr4$`Gene Symbol`)) #29822  suplication
expre4 <- limma::avereps(expr4[,-c(1,ncol(expr4))], ID = expr4$`Gene Symbol`)
expre4 <- as.data.frame(expre4)
expression4<- expre4[, -c(1,ncol(expre4))]
dim(expression4)
```
### use expr to filter pheno
```{r}
pheno4 <- phe4[rownames(phe4) %in% colnames(expression4), ]
```
### set group
```{r}
pheno4 <- pheno4 %>%
  mutate(
    Outcome = ifelse(`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1` == 'ABMR','ABMR',
                     ifelse(`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1`=='TCMR', 'TCMR', NA)))%>%
  filter(!is.na(Outcome))
table(pheno4$Outcome)
```
## 5. GSE98320
```{r}
gse5 <- 'GSE98320'
eSet5 <- getGEO('GSE98320')
eSet5 <- eSet5$GSE98320_series_matrix.txt.gz
exp5 <- exprs(eSet5)
phe5 <- pData(phenoData(eSet5))
feature5 <- fData(eSet5)
```
### clean feature
```{r}
f5 <-feature5%>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f5$'Gene Symbol' <- sapply(strsplit(f5$'Gene Symbol', ' /// '), function(x) x[1])
```
### probe sub (add gene sybol)
```{r}
gene_symbols5 <- f5[, c(1, 17)]
expr5 <- as.data.frame(exp5)
expr5[, 'ID'] <- rownames(expr5)
expr5 <- merge(x = expr5, y = gene_symbols5, by.x = 'ID')
```
### remove duplication (keep mean)
```{r}
table(duplicated(expr5$`Gene Symbol`)) #29911  suplication
expre5 <- limma::avereps(expr5[,-c(1,ncol(expr5))], ID = expr5$`Gene Symbol`)
expre5 <- as.data.frame(expre5)
expression5<- expre5[, -c(1,ncol(expre5))]
dim(expression5)
```
### use expr to filter pheno
```{r}
pheno5 <- phe5[rownames(phe5) %in% colnames(expression5), ]
```
### set group
```{r}
pheno5 <- pheno5 %>%
  mutate(
    Outcome = ifelse(`d96:ch1` == 'ABMR','ABMR',
                     ifelse(`d96:ch1`=='TCMR', 'TCMR', NA)))%>%
  filter(!is.na(Outcome))
table(pheno5$Outcome)
```
#group
```{r}
group <- c(pheno1$Outcome,pheno2$Outcome,pheno3$Outcome,pheno4$Outcome,pheno5$Outcome)
length(group)
```
# combine data sets
```{r,message=F}
library(sva)
common_genes <- Reduce(intersect, list(rownames(expression1), rownames(expression2), 
                                       rownames(expression3), rownames(expression4), 
                                       rownames(expression5)))
expression1_common <- expression1[common_genes, , drop = FALSE]
expression2_common <- expression2[common_genes, , drop = FALSE]
expression3_common <- expression3[common_genes, , drop = FALSE]
expression4_common <- expression4[common_genes, , drop = FALSE]
expression5_common <- expression5[common_genes, , drop = FALSE]
combined_expression <- cbind(expression1_common, expression2_common, 
                             expression3_common, expression4_common, 
                             expression5_common)
dim(combined_expression)
```
# checking batch effect (PCA plot)
```{r}
library(ggfortify)
gse_pca <- prcomp(t(combined_expression))
batches <- factor(rep(1:5, each = 10))

```

