---
title: "lab_2"
author: '520004915'
date: "2024-04-22"
output: html_document
---
---
title: '20.4'
author: '520004915'
date: "2024-04-19"
output: html_document
---
#load packages
```{r setup,s include=FALSE}
rm(list = ls())
library(caret)
library(cvTools)
library(dplyr)
library(e1071)
library(GEOquery)
library(ggplot2)
library(glmnet)
library(limma)
library(pheatmap)
library(randomForest)
library(stringr)
library(tidyverse)
```
#laod data sets & cleaning & gene probe sub
## 1. 'GSE36059'
```{r}
gse1 <- 'GSE36059'
eSet1 <- getGEO('GSE36059')
eSet1 <- eSet1$GSE36059_series_matrix.txt.gz
exp1 <- exprs(eSet1)
phe1 <- pData(phenoData(eSet1))
feature1 <- fData(eSet1)
```
### clean feature
```{r}
f1 <-feature1%>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
#f1[grepl('///', f1$`Gene Symbol`), ] after checking, i found they are different genes so i remove
f1$'Gene Symbol' <- sapply(strsplit(f1$'Gene Symbol', ' /// '), function(x) x[1])
```
### probe sub (add gene sybol)
```{r}
gene_symbols1 <- f1[, c(1, 11)]
expr1 <- as.data.frame(exp1)
expr1[, 'ID'] <- rownames(expr1)
expr1 <- merge(x = expr1, y = gene_symbols1, by.x = 'ID')
```
### add index
```{r}
#add index to phenotype
rownames(phe1) <- paste(rownames(phe1), '1', sep='_')
colnames(expr1) <- paste(colnames(expr1), '1', sep='_')
```
### remove duplication (keep mean)
```{r}
#table(duplicated(expr1$`Gene Symbol`)) #22902 suplication
expre1 <- limma::avereps(expr1[,-c(1,ncol(expr1))], ID = expr1$`Gene Symbol`)
expre1 <- as.data.frame(expre1)
expression1 <- expre1[, -c(1,ncol(expre1))]
dim(expression1)
```
### set group
```{r}
phe1 <- phe1[rownames(phe1) %in% colnames(expression1),]
phe1$Outcome <- ifelse(str_detect(phe1$title, 'non-rejecting'), 'Stable',
                       ifelse(str_detect(phe1$title, 'ABMR'), 'ABMR',
                              ifelse(str_detect(phe1$title, 'TCMR'), 'TCMR', 'Other')))
pheno1 <- phe1%>% filter(Outcome != 'Other')
table(pheno1$Outcome)
expression1 <- as.data.frame(expression1)
expression1 <- expression1[, colnames(expression1) %in% rownames(pheno1)]
expression1 <- as.matrix(expression1)
```
## 2. 'GSE129166'
```{r}
gse2 <- 'GSE129166'
eSet2 <- getGEO('GSE129166')
eSet2 <- eSet2$GSE129166_series_matrix.txt.gz
exp2 <- exprs(eSet2)
phe2 <- pData(phenoData(eSet2))
feature2 <- fData(eSet2)
```
### clean feature
```{r}
f2 <-feature2%>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f2$'Gene Symbol' <- sapply(strsplit(f2$'Gene Symbol', ' /// '), function(x) x[1])
```
### probe sub (add gene sybol)
```{r}
gene_symbols2 <- f2[, c(1, 11)]
expr2 <- as.data.frame(exp2)
expr2[, 'ID'] <- rownames(expr2)
expr2 <- merge(x = expr2, y = gene_symbols2, by.x = 'ID')
```
### add index
```{r}
#add index to phenotype
rownames(phe2) <- paste(rownames(phe2), '2', sep='_')
colnames(expr2) <- paste(colnames(expr2), '2', sep='_')
```
### remove duplication (keep mean)
```{r}
#table(duplicated(expr2$`Gene Symbol`)) #22902 suplication
expre2 <- limma::avereps(expr2[,-c(1,ncol(expr2))], ID = expr2$`Gene Symbol`)
expre2 <- as.data.frame(expre2)
expression2<- expre2[, -c(1,ncol(expre2))]
dim(expression2)
```
### set group
```{r}
phe2 <- phe2[rownames(phe2) %in% colnames(expression2),]
pheno2 <-phe2 %>% mutate(
  ABMR_Status = ifelse(`abmr (no:ch1` == '0_Yes:1): 1', 'ABMR', 'ABMR Stable'),
  TCMR_Status = ifelse(`tcmr (no:ch1` == '0_borderline:1_TCMR:2): 1', 'TCMR', 'TCMR Stable')
  )%>%
    filter(!(ABMR_Status == 'ABMR' & TCMR_Status == 'TCMR')) %>%
  mutate(
    Outcome = case_when(
      ABMR_Status == 'ABMR Stable' & TCMR_Status == 'TCMR Stable' ~ 'Stable',
      ABMR_Status == 'ABMR Stable' ~ 'TCMR',
      TCMR_Status == 'TCMR Stable' ~ 'ABMR',
      TRUE ~ NA_character_
    )
  )
table(pheno2$Outcome)
expression2 <- as.data.frame(expression2)
expression2 <- expression2[, colnames(expression2) %in% rownames(pheno2)]
expression2 <- as.matrix(expression2)
```
## 3. GSE192444
```{r}
gse3 <- 'GSE192444'
eSet3 <- getGEO('GSE192444')
eSet3 <- eSet3$GSE192444_series_matrix.txt.gz
exp3 <- exprs(eSet3)
phe3<- pData(phenoData(eSet3))
feature3 <- fData(eSet3)
```
### clean feature
```{r}
f3 <-feature3%>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f3$'Gene Symbol' <- sapply(strsplit(f3$'Gene Symbol', ' /// '), function(x) x[1])
```
### probe sub (add gene sybol)
```{r}
gene_symbols3 <- f3[, c(1, 15)]
expr3 <- as.data.frame(exp3)
expr3[, 'ID'] <- rownames(expr3)
expr3 <- merge(x = expr3, y = gene_symbols3, by.x = 'ID')
```
### add index
```{r}
#add index to phenotype
rownames(phe3) <- paste(rownames(phe3), '3', sep='_')
colnames(expr3) <- paste(colnames(expr3), '3', sep='_')
```
### remove duplication (keep mean)
```{r}
table(duplicated(expr3$`Gene Symbol`)) #29822  suplication
expre3 <- limma::avereps(expr3[,-c(1,ncol(expr3))], ID = expr3$`Gene Symbol`)
expre3 <- as.data.frame(expre3)
expression3<- expre3[, -c(1,ncol(expre3))]
dim(expression3)
```

### set group
```{r}
phe3 <- phe3[rownames(phe3) %in% colnames(expression3),]
phe3$Outcome <- ifelse(str_detect(phe3$`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1`, 'ABMR'), 'ABMR',
                       ifelse(str_detect(phe3$`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1`, 'TCMR'), 'TCMR',
                              ifelse(str_detect(phe3$`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1`, 'NR'), 'Stable', 'Other')))
pheno3 <- phe3%>% filter(Outcome != 'Other')
table(pheno3$Outcome)
expression3 <- as.data.frame(expression3)
expression3 <- expression3[, colnames(expression3) %in% rownames(pheno3)]
expression3 <- as.matrix(expression3)
```
## 4. GSE124203
```{r}
gse4 <- 'GSE124203'
eSet4 <- getGEO('GSE124203')
eSet4 <- eSet4$GSE124203_series_matrix.txt.gz
exp4 <- exprs(eSet4)
phe4<- pData(phenoData(eSet4))
feature4 <- fData(eSet4)
```
### clean feature
```{r}
f4 <-feature4%>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f4$'Gene Symbol' <- sapply(strsplit(f4$'Gene Symbol', ' /// '), function(x) x[1])
```
### probe sub (add gene sybol)
```{r}
gene_symbols4 <- f4[, c(1, 15)]
expr4 <- as.data.frame(exp4)
expr4[, 'ID'] <- rownames(expr4)
expr4 <- merge(x = expr4, y = gene_symbols4, by.x = 'ID')
```
### add index
```{r}
#add index to phenotype
rownames(phe4) <- paste(rownames(phe4), '4', sep='_')
colnames(expr4) <- paste(colnames(expr4), '4', sep='_')
```
### remove duplication (keep mean)
```{r}
table(duplicated(expr4$`Gene Symbol`)) #29822  suplication
expre4 <- limma::avereps(expr4[,-c(1,ncol(expr4))], ID = expr4$`Gene Symbol`)
expre4 <- as.data.frame(expre4)
expression4<- expre4[, -c(1,ncol(expre4))]
dim(expression4)
```
### set group
```{r}
phe4 <- phe4[rownames(phe4) %in% colnames(expression4),]
phe4$Outcome <- ifelse(str_detect(phe4$`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1`, 'ABMR'), 'ABMR',
                       ifelse(str_detect(phe4$`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1`, 'TCMR'), 'TCMR',
                              ifelse(str_detect(phe4$`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1`, 'NR'), 'Stable', 'Other')))
pheno4 <- phe4%>% filter(Outcome != 'Other')
table(pheno4$Outcome)
expression4 <- as.data.frame(expression4)
expression4 <- expression4[, colnames(expression4) %in% rownames(pheno4)]
expression4 <- as.matrix(expression4)
```
## 5. GSE98320
```{r}
gse5 <- 'GSE98320'
eSet5 <- getGEO('GSE98320')
eSet5 <- eSet5$GSE98320_series_matrix.txt.gz
exp5 <- exprs(eSet5)
phe5 <- pData(phenoData(eSet5))
feature5 <- fData(eSet5)
```
### clean feature
```{r}
f5 <-feature5%>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f5$'Gene Symbol' <- sapply(strsplit(f5$'Gene Symbol', ' /// '), function(x) x[1])
```
### probe sub (add gene sybol)
```{r}
gene_symbols5 <- f5[, c(1, 17)]
expr5 <- as.data.frame(exp5)
expr5[, 'ID'] <- rownames(expr5)
expr5 <- merge(x = expr5, y = gene_symbols5, by.x = 'ID')
```
### add index
```{r}
#add index to phenotype
rownames(phe5) <- paste(rownames(phe5), '5', sep='_')
colnames(expr5) <- paste(colnames(expr5), '5', sep='_')
```
### remove duplication (keep mean)
```{r}
table(duplicated(expr5$`Gene Symbol`)) #29911  suplication
expre5 <- limma::avereps(expr5[,-c(1,ncol(expr5))], ID = expr5$`Gene Symbol`)
expre5 <- as.data.frame(expre5)
expression5<- expre5[, -c(1,ncol(expre5))]
dim(expression5)
```
### set group
```{r}
phe5 <- phe5[rownames(phe5) %in% colnames(expression5),]
pheno5 <- phe5 %>%
  mutate(
    Outcome = ifelse(`d96:ch1` == 'ABMR','ABMR',
                     ifelse(`d96:ch1`=='TCMR', 'TCMR', NA)))%>%
  filter(!is.na(Outcome))
table(pheno5$Outcome)
expression5 <- as.data.frame(expression5)
expression5 <- expression5[, colnames(expression5) %in% rownames(pheno5)]
expression5 <- as.matrix(expression5)
```

#alignment
```{r}
top_75_percentile <- function(eMat) {
  eMat <- as.data.frame(eMat)
  hist(rowSums(eMat),breaks=50,main='row sum dist')
  threshold <-quantile(rowSums(eMat),0.25)
  filtered_eMat <- eMat[rowSums(eMat) > threshold,]
  return (filtered_eMat)
}
expression1 <- top_75_percentile(expression1)
expression2 <- top_75_percentile(expression2)
expression3 <- top_75_percentile(expression3)
expression4 <- top_75_percentile(expression4)
expression5 <- top_75_percentile(expression5)
phen1 <- phe1[rownames(phe1) %in% colnames(expression1),]
phen2 <- phe2[rownames(phe2) %in% colnames(expression2),]
phen3 <- phe3[rownames(phe3) %in% colnames(expression3),]
phen4 <- phe4[rownames(phe4) %in% colnames(expression4),]
phen5 <- phe5[rownames(phe5) %in% colnames(expression5),]
```


# combine data sets
```{r,message=F}
common_genes <- Reduce(intersect, list(rownames(expression1), rownames(expression2), 
                                       rownames(expression3), rownames(expression4), 
                                       rownames(expression5)))
expression1_common <- expression1[common_genes, ]
expression2_common <- expression2[common_genes, ]
expression3_common <- expression3[common_genes, ]
expression4_common <- expression4[common_genes, ]
expression5_common <- expression5[common_genes, ]
combined_expression <- cbind(expression1_common, expression2_common, 
                             expression3_common, expression4_common, 
                             expression5_common)
dim(combined_expression)
```
# normalise
```{r}
eMat <- normalizeBetweenArrays(combined_expression)
```

# pca checking batch effect
```{r}
gse_pca <- prcomp(t(eMat))
group <- c(pheno1$Outcome,pheno2$Outcome,pheno3$Outcome,pheno4$Outcome,pheno5$Outcome)
batch_info <- c(rep(1, ncol(expression1_common)),
                rep(2, ncol(expression2_common)),
                rep(3, ncol(expression3_common)),
                rep(4, ncol(expression4_common)),
                rep(5, ncol(expression5_common)))
df_toplot <- data.frame(group,
                        pc1 = gse_pca$x[,1], pc2 = gse_pca$x[,2],
                        batch = batch_info)
df_toplot$batch <- as.factor(df_toplot$batch)
ggplot(df_toplot, aes(x = pc1, y = pc2, color = group)) + 
  geom_point(size = 1.5) + labs(colour = "Outcome") +
  theme_minimal()
ggplot(df_toplot, aes(x = pc1, y = pc2, color = batch)) + 
  geom_point(size = 1.5) + labs(colour = "Batch") +
  theme_minimal()
```
# remove batch effect
```{r}
library(sva)
batch <- factor(c(rep(gse1, ncol(expression1_common)), 
                  rep(gse2, ncol(expression2_common)), 
                  rep(gse3, ncol(expression3_common)),
                  rep(gse4, ncol(expression4_common)),
                  rep(gse4, ncol(expression5_common))))
mod <- model.matrix(~1, data=data.frame(ones=rep(1, 2768)))
combat_eMat <- sva::ComBat(dat=eMat, batch=batch, mod=mod, par.prior=TRUE, prior.plots=FALSE)
dim(combat_eMat)
```
# PCA check result from batch effect
```{r}
gse_pca <- prcomp(t(combat_eMat))
df_toplot <- data.frame(group,
                        pc1 = gse_pca$x[,1], pc2 = gse_pca$x[,2],
                        batch = batch_info)
df_toplot$batch <- as.factor(df_toplot$batch)
ggplot(df_toplot, aes(x = pc1, y = pc2, color = group)) + 
  geom_point(size = 1.5) + labs(colour = "Outcome") +
  theme_minimal()
ggplot(df_toplot, aes(x = pc1, y = pc2, color = batch)) + 
  geom_point(size = 1.5) + labs(colour = "Batch") +
  theme_minimal()
```
#set general group
```{r}
phen1 <- pheno1[rownames(pheno1) %in% colnames(expression1_common),]
phen2 <- pheno2[rownames(pheno2) %in% colnames(expression2_common),]
phen3 <- pheno3[rownames(pheno3) %in% colnames(expression3_common),]
phen4 <- pheno4[rownames(pheno4) %in% colnames(expression4_common),]
phen5 <- pheno5[rownames(pheno5) %in% colnames(expression5_common),]
group <- c(phen1$Outcome,phen2$Outcome,phen3$Outcome,phen4$Outcome,phen5$Outcome)
length(group)
```
# adjust expression
```{r}
exprSet_corrected <- combat_eMat[, colnames(combat_eMat) %in% c(rownames(phen1), rownames(phen2),rownames(phen3),rownames(phen4),rownames(phen5))]
dim(exprSet_corrected)
```
#limma selection
```{r}
group_factor <- factor(group)
design <- model.matrix(~ 0 + group_factor)
colnames(design) <- levels(group_factor)
rownames(design) <- colnames(exprSet_corrected)
contrasts_matrix <- makeContrasts(
  Stable_vs_ABMR = Stable - ABMR,
  Stable_vs_TCMR =Stable - TCMR,
  ABRM_vs_TCMR = Stable - ((ABMR+TCMR)*(1/2)),
  levels = design
)
fit <- lmFit(exprSet_corrected, design)
fit <- contrasts.fit(fit, contrasts = contrasts_matrix)
fit <- eBayes(fit)
tT <-topTable(fit,number=Inf,adjust ='BH')%>% arrange('adj.P.Val')
top100 <- rownames(tT[1:100,])
top100
```
#then lasso selection
```{r}
set.seed(3888)
genes <- exprSet_corrected[top100,]
group_numeric <- as.numeric(factor(group))
x<- t(genes)
cv_lasso <- cv.glmnet(x, group_numeric, alpha = 1)
opt_lambda <- cv_lasso$lambda.1se
final_lasso <- glmnet(x, group_numeric, alpha = 1, lambda = opt_lambda)
lasso_coef <- coef(final_lasso, s = opt_lambda)
selected_genes <- rownames(genes)[which(lasso_coef[-1,] != 0)]
selected_genes #38
```
# heatmap
```{r}
normalized_expression <- scale(exprSet_corrected[selected_genes, ])
pheatmap(normalized_expression, 
         scale = "row",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete", 
         main = 'Heatmap of Selected Genes',
         fontsize = 5
)
```

```{r}
set.seed(3888)

X = t(exprSet_corrected)
y = factor(group)

cvK = 5
n_sim = 20

cv_accuracy_rf = cv_accuracy_svm = cv_accuracy_knn = numeric(n_sim)

for (i in 1:n_sim) {
  cvSets = cvFolds(nrow(X), cvK) 
  cv_accuracy_folds_rf = cv_accuracy_folds_svm = cv_accuracy_folds_knn = numeric(cvK)

  for (j in 1:cvK) {
    test_id = cvSets$subsets[cvSets$which == j]
    X_train = X[-test_id,]
    X_test = X[test_id,]
    y_train = y[-test_id]
    y_test = y[test_id]
    
    design <- model.matrix(~ 0 + y_train)
    colnames(design) <- levels(y_train)
    contrasts_matrix <- makeContrasts(
      Stable_vs_ABMR = Stable - ABMR,
      Stable_vs_TCMR =Stable - TCMR,
      ABRM_vs_TCMR = Stable - ((ABMR+TCMR)*(1/2)),
      levels = design
    )
    
    fit <- lmFit(t(X_train), design)
    fit <- contrasts.fit(fit, contrasts = contrasts_matrix)
    fit <- eBayes(fit)
    tT <- topTable(fit, number = 50) %>% arrange('adj.P.Val')
    
    X_train <- X_train[, rownames(tT)]
    X_test <- X_test[, rownames(tT)]
    
    group_numeric <- as.numeric(y_train)
    cv_lasso <- cv.glmnet(X_train, group_numeric, alpha = 1)
    opt_lambda <- cv_lasso$lambda.1se
    final_lasso <- glmnet(X_train, group_numeric, alpha = 1, lambda = opt_lambda)
    lasso_coef <- coef(final_lasso, s = opt_lambda)
    lasso_genes <- colnames(X_train)[which(lasso_coef[-1,] != 0)]

    X_train = X_train[,lasso_genes]
    X_test = X_test[,lasso_genes]
    
    rf_fit = randomForest(x = X_train, y = y_train)
    predictions_rf = predict(rf_fit, X_test)
    cv_accuracy_folds_rf[j] = mean(y_test == predictions_rf)
    
    knn_fit = class::knn(train = X_train, test = X_test, cl = y_train, k = 5)
    cv_accuracy_folds_knn[j] = mean(y_test == knn_fit)

    svm_fit = svm(X_train, y_train)
    predictions_svm = predict(svm_fit, X_test)
    cv_accuracy_folds_svm[j] = mean(y_test == predictions_svm)
  }
  
  cv_accuracy_rf[i] = mean(cv_accuracy_folds_rf)
  cv_accuracy_knn[i] = mean(cv_accuracy_folds_knn)
  cv_accuracy_svm[i] = mean(cv_accuracy_folds_svm)
}

boxplot(cv_accuracy_rf, cv_accuracy_knn, cv_accuracy_svm, names = c("RF", "KNN (k = 5)", "SVM"),
        xlab = "Classifier", ylab = "Accuracy", main = "Performance of classifiers on top DE genes using Limma-Lasso")
```

