---
title: "FIRST_5_STEPS"
author: '520004915'
date: "2024-04-11"
output: html_document
---
```{r setup, include=FALSE}
library(caret)
library(cvTools)
library(dplyr)
library(e1071)
library(GEOquery)
library(ggplot2)
library(glmnet)
library(limma)
library(pheatmap)
library(randomForest)
library(stringr)
library(tidyverse)

rm(list = ls())
```


Loading in the datasets

```{r}
gse1 ='GSE36059'
gse2 ='GSE129166'

eSet1 <- getGEO('GSE36059')
eSet1 <- eSet1$GSE36059_series_matrix.txt.gz
eSet2 <- getGEO('GSE129166')
eSet2 <- eSet2$GSE129166_series_matrix.txt.gz

exp1 <- exprs(eSet1)
phe1 <- pData(phenoData(eSet1))
feature1 <- fData(eSet1)

exp2 <- exprs(eSet2)
phe2 <- pData(phenoData(eSet2))
feature2 <- fData(eSet2)
```


Cleaning and merging of the datasets

```{r}
set.seed(3888)

#clean features
f1 <-feature1 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f1$'Gene Symbol' <- sapply(strsplit(f1$'Gene Symbol', ' /// '), function(x) x[1])
f1$'Gene Symbol' <- make.unique(f1$'Gene Symbol', sep = '_')

f2 <-feature2 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f2$'Gene Symbol' <- sapply(strsplit(f2$'Gene Symbol', ' /// '), function(x) x[1])
f2$'Gene Symbol' <- make.unique(f2$'Gene Symbol', sep = '_')

## Rename probe IDs in expression matrix to gene symbols
exp1 <- as.data.frame(exp1)
# Probe ID in column 1, gene symbol in column 11
gene_symbols1 <- f1[, c(1, 11)]
exp1[, 'ID'] <- rownames(exp1)
exp1 <- merge(x = exp1, y = gene_symbols1, by.x = 'ID')
f1 <- f1 %>% filter(`ID` %in% exp1$`ID`)
exp1[, 'Gene Symbol'] <- make.unique(exp1[, 'Gene Symbol'], sep = '_')
# Set gene symbols as the rownames of the data frame
rownames(exp1) <- exp1[, 'Gene Symbol']
exp1[, c('ID', 'Gene Symbol')] <- NULL
exp1 <- as.matrix(exp1)

exp2 <- as.data.frame(exp2)
gene_symbols2 <- f2[, c(1, 11)]
exp2[, 'ID'] <- rownames(exp2)
exp2 <- merge(x = exp2, y = gene_symbols2, by.x = 'ID')
f2 <- f2 %>% filter(`ID` %in% exp2$`ID`)
exp2[, 'Gene Symbol'] <- make.unique(exp2[, 'Gene Symbol'], sep = '_')
rownames(exp2) <- exp2[, 'Gene Symbol']
exp2[, c('ID', 'Gene Symbol')] <- NULL
exp2 <- as.matrix(exp2)

#conbime gene expressions
table(rownames(exp1) %in% rownames(exp2))
exp1 <- exp1[intersect(rownames(exp1),rownames(exp2)),]
exp2 <- exp2[intersect(rownames(exp1),rownames(exp2)),]

if(!identical(rownames(phe1),colnames(exp1))) exp1 = exp1[,match(rownames(phe1),colnames(exp1))]
if(!identical(rownames(phe2),colnames(exp2))) exp2 = exp2[,match(rownames(phe2),colnames(exp2))]

exp <- cbind(exp1,exp2)

#remove batch effects
batch_info <- c(rep(1, ncol(exp1)), rep(2, ncol(exp2)))
exprSet_corrected <- removeBatchEffect(exp, batch_info)
##set groups
phe1$outcome <- ifelse(str_detect(phe1$title, 'non-rejecting'), 'non-rejection',
                       ifelse(str_detect(phe1$title, 'ABMR'), 'ABMR',
                              ifelse(str_detect(phe1$title, 'TCMR'), 'TCMR', 'Other')))
phe1 <- phe1%>% filter(outcome != 'Other')#remove other

phe2 <-phe2 %>% mutate(
  ABMR_Status = ifelse(`abmr (no:ch1` == '0_Yes:1): 1', 'ABMR', 'ABMR Stable'),
  TCMR_Status = ifelse(`tcmr (no:ch1` == '0_borderline:1_TCMR:2): 1', 'TCMR', 'TCMR Stable')
  )%>%
    filter(!(ABMR_Status == 'ABMR' & TCMR_Status == 'TCMR')) %>%
  mutate(
    outcome = case_when(
      ABMR_Status == 'ABMR Stable' & TCMR_Status == 'TCMR Stable' ~ 'non-rejection',
      ABMR_Status == 'ABMR Stable' ~ 'TCMR',
      TCMR_Status == 'TCMR Stable' ~ 'ABMR',
      TRUE ~ NA_character_
    )
  )
###remove both TCMR and ABMR, then if the original sting was TCMR Stable, then it appears to be ABMR

## Remove corresponding columns of geo accession data from expression matrix that were dropped in pheno data
exprSet_corrected <- exprSet_corrected[, colnames(exprSet_corrected) %in% c(phe1$geo_accession, phe2$geo_accession)]

group <- table(c(phe1$outcome, phe2$outcome))

#normalisation
combine_symbol <- union(f1$`Gene Symbol`, f2$`Gene Symbol`)
eMat <- exprSet_corrected[rownames(exprSet_corrected) %in% combine_symbol, ]
gene_means <- rowMeans(eMat)
gene_percentile <- rank(gene_means) / length(gene_means)
threshold <- 0.25
filtered_genes <- eMat[gene_percentile > threshold,]
## [normalisation]
exprSet <- normalizeBetweenArrays(filtered_genes)
qx <- as.numeric(quantile(exprSet,c(0.,0.25,0.5,0.75,0.99,1.0),na.rm=T))
logC <- (qx[5]>100) ||
           (qx[6]-qx[1]>50 && qx[2]>0) ||
            (qx[2]>0 && qx[2]<1 && qx[4]>1 && qx[4]<2)
if (logC){
  exprSet [which(exprSet <=0)] <- NaN
  exprSet_clean <- log2(exprSet)
  print('log2 transformation finished')
}else{
  print('does not need log2 transformation')
}
#normalisaton done, see how it looks
boxplot(exp)#after combination
boxplot(exprSet_corrected)#bacth effected
boxplot(exprSet)#normaled
```


Exploring the dataset with PCA

```{r}
set.seed(3888)

gse_pca <- prcomp(t(exprSet))
df_toplot <- data.frame(c(phe1$outcome, phe2$outcome),
                        pc1 = gse_pca$x[,1], pc2 = gse_pca$x[,2])

ggplot(df_toplot, aes(x = pc1, y = pc2, color = c(phe1$outcome, phe2$outcome))) + 
  geom_point(size = 1.5) + labs(colour = "Outcome") +
  theme_minimal()
```


Selecting the top differentially expressed genes using Limma

```{r}
set.seed(3888)

combined_outcome = c(phe1$outcome, phe2$outcome)
design_matrix <- model.matrix(~ 0 + combined_outcome)
colnames(design_matrix) <- c("ABMR", "non_rejection", "TCMR")

contrast_matrix <- makeContrasts(
    nonrejVSabmr = non_rejection - ABMR,
    nonrejVStcmr = non_rejection - TCMR,
    nonrejVSrej = non_rejection - (1 / 2 * (ABMR + TCMR)), levels = colnames(design_matrix))

limma_fit <- lmFit(exprSet, design = design_matrix)
limma_fit <- contrasts.fit(limma_fit, contrasts = contrast_matrix)
limma_fit <- eBayes(limma_fit)
tT <- topTable(limma_fit, n = nrow(limma_fit))

top10 <- rownames(tT[order(tT$adj.P.Val),][1:10,])
top50 <- rownames(tT[order(tT$adj.P.Val),][1:50,])
top100 <- rownames(tT[order(tT$adj.P.Val),][1:100,])
```

```{r}
topTable(limma_fit, n = nrow(limma_fit), coef = "nonrejVSabmr")
topTable(limma_fit, n = nrow(limma_fit), coef = "nonrejVStcmr")
```


Visualising features that are different between stable and rejection with volcano plot

```{r}
ggplot(tT, aes(nonrejVSabmr, -log10(P.Value))) +
  geom_point(aes(colour = -log10(P.Value)), alpha = 1/3, size = 1) +
  scale_colour_gradient(low = "blue", high = "red") +
  xlab("log2 fold change") + ylab("-log10 p-value") + ggtitle("Volcano plot for Non-rejection vs ABMR")

ggplot(tT, aes(nonrejVStcmr, -log10(P.Value))) +
  geom_point(aes(colour = -log10(P.Value)), alpha = 1/3, size = 1) +
  scale_colour_gradient(low = "blue", high = "red") +
  xlab("log2 fold change") + ylab("-log10 p-value") + ggtitle("Volcano plot for Non-rejection vs TCMR")

ggplot(tT, aes(nonrejVSrej, -log10(P.Value))) +
  geom_point(aes(colour = -log10(P.Value)), alpha = 1/3, size = 1) +
  scale_colour_gradient(low = "blue", high = "red") +
  xlab("log2 fold change") + ylab("-log10 p-value") + ggtitle("Volcano plot for Non-rejection vs Rejection")
```


Visualising the heatmap for the top 100 differentially expressed genes selected using Limma

```{r, message=F}
top100_expr <- exprSet[rownames(exprSet) %in% top100, ]
breaks <- seq(min(top100_expr), max(top100_expr), length.out = 100)
color_palette <- colorRampPalette(c('dodgerblue4', 'white', 'yellow'))(99)

sample_annotation <- data.frame(Group = factor(combined_outcome))
rownames(sample_annotation) <- colnames(top100_expr)

annotation_colors <- c(ABMR='firebrick4', TCMR='darkseagreen3', `non-rejection`='white')

annotation_col_colors <- list(Group = annotation_colors)

pheatmap(top100_expr,
         color = color_palette,
         breaks = breaks,
         annotation_col = sample_annotation,
         annotation_colors = annotation_col_colors,
         fontsize = 3,
         fontsize_row = 3,
         fontsize_col = 3,
         clustering_method = "complete",
         scale = "row",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         legend = TRUE)
```


Selecting the top differentially expressed genes using Lasso

```{r, message=F}
group_numeric <- as.numeric(factor(combined_outcome))
x <- t(exprSet)
set.seed(3888)
cv_lasso <- cv.glmnet(x, group_numeric, alpha = 1)
opt_lambda <- cv_lasso$lambda.min

final_lasso <- glmnet(x, group_numeric, alpha = 1, lambda = opt_lambda)
lasso_coef <- coef(final_lasso, s = opt_lambda)
selected_genes <- rownames(exprSet)[which(lasso_coef[-1,] != 0)]
```


Visualising the heatmap for the top differentially expressed genes selected using Lasso

```{r}
selected_lasso <- exprSet[rownames(exprSet) %in% selected_genes, ]
breaks <- seq(min(selected_lasso), max(selected_lasso), length.out = 136)
color_palette <- colorRampPalette(c('dodgerblue4', 'white', 'yellow'))(135)

sample_annotation <- data.frame(Group = factor(combined_outcome))
rownames(sample_annotation) <- colnames(selected_lasso)

annotation_colors <- c(ABMR='firebrick4', TCMR='darkseagreen3', `non-rejection`='white')

annotation_col_colors <- list(Group = annotation_colors)

pheatmap(selected_lasso,
         color = color_palette,
         breaks = breaks,
         annotation_col = sample_annotation,
         annotation_colors = annotation_col_colors,
         fontsize = 3,
         fontsize_row = 3,
         fontsize_col = 3,
         clustering_method = "complete",
         scale = "row",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         legend = TRUE)
```


Evaluating the performance of classifiers on unseen data with feature selection using Limma

```{r}
set.seed(3888)

X = t(exprSet)
y = factor(combined_outcome)

cvK = 5   
n_sim = 20

cv_accuracy10_rf = cv_accuracy50_rf = cv_accuracy100_rf = numeric(n_sim)
cv_accuracy10_knn = cv_accuracy50_knn = cv_accuracy100_knn = numeric(n_sim)
cv_accuracy10_svm = cv_accuracy50_svm = cv_accuracy100_svm = numeric(n_sim)

for (i in 1:n_sim) {
  cvSets = cvFolds(nrow(X), cvK) 
  cv_accuracy_folds10_rf = cv_accuracy_folds50_rf = cv_accuracy_folds100_rf = numeric(cvK)
  cv_accuracy_folds10_knn = cv_accuracy_folds50_knn = cv_accuracy_folds100_knn = numeric(cvK)
  cv_accuracy_folds10_svm = cv_accuracy_folds50_svm = cv_accuracy_folds100_svm = numeric(cvK)

  for (j in 1:cvK) {
    test_id = cvSets$subsets[cvSets$which == j]
    X_train = X[-test_id,]
    X_test = X[test_id,]
    y_train = y[-test_id]
    y_test = y[test_id]
    
    design_matrix <- model.matrix(~ 0 + y_train)
    colnames(design_matrix) <- c("ABMR", "non_rejection", "TCMR")

    contrast_matrix <- makeContrasts(
      nonrejVSabmr = non_rejection - ABMR,
      nonrejVStcmr = non_rejection - TCMR,
      nonrejVSrej = non_rejection - (1 / 2 * (ABMR + TCMR)), levels = colnames(design_matrix))

    limma_fit <- lmFit(t(X_train), design = design_matrix)
    limma_fit <- contrasts.fit(limma_fit, contrasts = contrast_matrix)
    limma_fit <- eBayes(limma_fit)
    limma_top <- topTable(limma_fit, n = 100)
    
    DE_genes10 <- rownames(limma_top)[1:10]
    DE_genes50 <- rownames(limma_top)[1:50]
    DE_genes100 <- rownames(limma_top)[1:100]
    
    X_train10 = X_train[,DE_genes10]
    X_test10 = X_test[,DE_genes10]
    
    X_train50 = X_train[,DE_genes50]
    X_test50 = X_test[,DE_genes50]
    
    X_train100 = X_train[,DE_genes100]
    X_test100 = X_test[,DE_genes100]
    
    # Random Forest
    rf_fit10 = randomForest(x = X_train10, y = y_train)
    predictions_rf10 = predict(rf_fit10, X_test10)
    cv_accuracy_folds10_rf[j] = mean(y_test == predictions_rf10)
    
    rf_fit50 = randomForest(x = X_train50, y = y_train)
    predictions_rf50 = predict(rf_fit50, X_test50)
    cv_accuracy_folds50_rf[j] = mean(y_test == predictions_rf50)
    
    rf_fit100 = randomForest(x = X_train100, y = y_train)
    predictions_rf100 = predict(rf_fit100, X_test100)
    cv_accuracy_folds100_rf[j] = mean(y_test == predictions_rf100)
    
    # KNN (k = 5)
    knn_fit10 = class::knn(train = X_train10, test = X_test10, cl = y_train, k = 5)
    cv_accuracy_folds10_knn[j] = mean(y_test == knn_fit10)
    
    knn_fit50 = class::knn(train = X_train50, test = X_test50, cl = y_train, k = 5)
    cv_accuracy_folds50_knn[j] = mean(y_test == knn_fit50)
    
    knn_fit100 = class::knn(train = X_train100, test = X_test100, cl = y_train, k = 5)
    cv_accuracy_folds100_knn[j] = mean(y_test == knn_fit100)
    
    # SVM
    svm_fit10 = svm(X_train10, y_train)
    predictions_svm10 = predict(svm_fit10, X_test10)
    cv_accuracy_folds10_svm[j] = mean(y_test == predictions_svm10)
    
    svm_fit50 = svm(X_train50, y_train)
    predictions_svm50 = predict(svm_fit50, X_test50)
    cv_accuracy_folds50_svm[j] = mean(y_test == predictions_svm50)
    
    svm_fit100 = svm(X_train100, y_train)
    predictions_svm100 = predict(svm_fit100, X_test100)
    cv_accuracy_folds100_svm[j] = mean(y_test == predictions_svm100)
  }
  
  cv_accuracy10_rf[i] = mean(cv_accuracy_folds10_rf)
  cv_accuracy50_rf[i] = mean(cv_accuracy_folds50_rf)
  cv_accuracy100_rf[i] = mean(cv_accuracy_folds100_rf)
  
  cv_accuracy10_knn[i] = mean(cv_accuracy_folds10_knn)
  cv_accuracy50_knn[i] = mean(cv_accuracy_folds50_knn)
  cv_accuracy100_knn[i] = mean(cv_accuracy_folds100_knn)

  cv_accuracy10_svm[i] = mean(cv_accuracy_folds10_svm)
  cv_accuracy50_svm[i] = mean(cv_accuracy_folds50_svm)
  cv_accuracy100_svm[i] = mean(cv_accuracy_folds100_svm)
}

# boxplot(cv_accuracy10_rf, cv_accuracy10_knn, cv_accuracy10_svm, names = c("RF", "KNN (k = 5)", "SVM"),
#         xlab = "Classifier", ylab = "Accuracy", main = "Performance of classifiers on top 10 DE genes using Limma")
# boxplot(cv_accuracy50_rf, cv_accuracy50_knn, cv_accuracy50_svm, names = c("RF", "KNN (k = 5)", "SVM"),
#         xlab = "Classifier", ylab = "Accuracy", main = "Performance of classifiers on top 50 DE genes using Limma")
# boxplot(cv_accuracy100_rf, cv_accuracy100_knn, cv_accuracy100_svm, names = c("RF", "KNN (k = 5)", "SVM"),
#         xlab = "Classifier", ylab = "Accuracy", main = "Performance of classifiers on top 100 DE genes using Limma")

boxplot(cv_accuracy10_rf, cv_accuracy50_rf, cv_accuracy100_rf,
        cv_accuracy10_knn, cv_accuracy50_knn, cv_accuracy100_knn,
        cv_accuracy10_svm, cv_accuracy50_svm, cv_accuracy100_svm,
        names = c("RF - 10", "RF - 50", "RF - 100", "KNN - 10", "KNN - 50", "KNN - 100", "SVM - 10", "SVM - 50", "SVM - 100"),
        cex.axis = 0.65)
```


Evaluating the performance of classifiers on unseen data with feature selection using Lasso

```{r}
set.seed(3888)

X = t(exprSet)
y = factor(combined_outcome)

cvK = 5   
n_sim = 20

cv_accuracy_rf = cv_accuracy_svm = cv_accuracy_knn = numeric(n_sim)

for (i in 1:n_sim) {
  cvSets = cvFolds(nrow(X), cvK) 
  cv_accuracy_folds_rf = cv_accuracy_folds_svm = cv_accuracy_folds_knn = numeric(cvK)

  for (j in 1:cvK) {
    test_id = cvSets$subsets[cvSets$which == j]
    X_train = X[-test_id,]
    X_test = X[test_id,]
    y_train = y[-test_id]
    y_test = y[test_id]
    
    group_numeric <- as.numeric(y_train)
    cv_lasso <- cv.glmnet(X_train, group_numeric, alpha = 1)
    opt_lambda <- cv_lasso$lambda.min
    
    final_lasso <- glmnet(X_train, group_numeric, alpha = 1, lambda = opt_lambda)
    lasso_coef <- coef(final_lasso, s = opt_lambda)
    lasso_genes <- rownames(exprSet)[which(lasso_coef[-1,] != 0)]

    X_train = X_train[,lasso_genes]
    X_test = X_test[,lasso_genes]
    
    rf_fit = randomForest(x = X_train, y = y_train)
    predictions_rf = predict(rf_fit, X_test)
    cv_accuracy_folds_rf[j] = mean(y_test == predictions_rf)
    
    knn_fit = class::knn(train = X_train, test = X_test, cl = y_train, k = 5)
    cv_accuracy_folds_knn[j] = mean(y_test == knn_fit)

    svm_fit = svm(X_train, y_train)
    predictions_svm = predict(svm_fit, X_test)
    cv_accuracy_folds_svm[j] = mean(y_test == predictions_svm)
  }
  
  cv_accuracy_rf[i] = mean(cv_accuracy_folds_rf)
  cv_accuracy_knn[i] = mean(cv_accuracy_folds_knn)
  cv_accuracy_svm[i] = mean(cv_accuracy_folds_svm)
}

boxplot(cv_accuracy_rf, cv_accuracy_knn, cv_accuracy_svm, names = c("RF", "KNN (k = 5)", "SVM"),
        xlab = "Classifier", ylab = "Accuracy", main = "Performance of classifiers on top DE genes using Lasso")
```

