---
title: "DATA3888 Kidney 20: Code"
author: 'Kidney_20'
output:
  html_document:
    fig_caption: yes
    self_contained: yes
    theme: flatly
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
---

## Loading Packages

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(error = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(tidy = TRUE)
library(caret)
library(ComplexUpset)
library(cvTools)
library(dplyr)
library(e1071)
library(GEOquery)
library(ggplot2)
library(glmnet)
library(grid)
library(gridExtra)
library(limma)
library(pheatmap)
library(plyr)
library(pROC)
library(precrec)
library(RColorBrewer)
library(randomForest)
library(stringr)
library(sva)
library(tidyverse)
library(UpSetR)
```

## Data Processing

### 1. GSE36059

```{r}
#loading dataset1 GSE36059
gse1 <- 'GSE36059'
eSet1 <- getGEO('GSE36059')
eSet1 <- eSet1$GSE36059_series_matrix.txt.gz
exp1 <- exprs(eSet1)
phe1 <- pData(eSet1)
feature1 <- fData(eSet1)
```

#### Cleaning feature data

```{r}
#remove missing values
f1 <- feature1 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
#keep the first gene
f1$'Gene Symbol' <- sapply(strsplit(f1$'Gene Symbol', ' /// '), function(x) x[1])
```

#### Substituting Gene Probe ID with Gene Symbol

```{r}
#gene probe -> gene symbol
gene_symbols1 <- f1[, c(1, 11)]
expr1 <- as.data.frame(exp1)
expr1[, 'ID'] <- rownames(expr1)
expr1 <- merge(x = expr1, y = gene_symbols1, by.x = 'ID')
```

#### Adding Index

```{r}
#_1 i where i is the index of data sets
rownames(phe1) <- paste(rownames(phe1), '1', sep='_')
colnames(expr1) <- paste(colnames(expr1), '1', sep='_')
```

#### Removing Duplicated Gene Symbols - Keeping the Mean Gene Expression

```{r}
#Keeping the Mean Gene Expression
expre1 <- limma::avereps(expr1[, -c(1, ncol(expr1))], ID = expr1$`Gene Symbol`)
expression1 <- as.data.frame(expre1)
```

#### Filtering of Outcome Groups

```{r}
phe1$Outcome <- ifelse(str_detect(phe1$title, 'non-rejecting'), 'Stable',
                       ifelse(str_detect(phe1$title, 'ABMR'), 'ABMR',
                              ifelse(str_detect(phe1$title, 'TCMR'), 'TCMR', 'Other')))
pheno1 <- phe1 %>% filter(Outcome != 'Other')

expression1 <- as.data.frame(expression1)
#alignment
expression1 <- expression1[, colnames(expression1) %in% rownames(pheno1)]
expression1 <- as.matrix(expression1)
```

### 2. GSE129166

```{r}
#loading dataset2 GSE36059
gse2 <- 'GSE129166'
eSet2 <- getGEO('GSE129166')
eSet2 <- eSet2$GSE129166_series_matrix.txt.gz
exp2 <- exprs(eSet2)
phe2 <- pData(eSet2)
feature2 <- fData(eSet2)
```

#### Filtering to Biopsy Data Only

```{r}
#Filtering to biopsy data
phe2 <- phe2 %>% filter(str_detect(title, 'BIOPSY'))
exp2 <- exp2[, rownames(phe2)]
```

#### Cleaning Feature Data

```{r}
f2 <- feature2 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f2$'Gene Symbol' <- sapply(strsplit(f2$'Gene Symbol', ' /// '), function(x) x[1])
```

#### Substituting Gene Probe ID with Gene Symbol

```{r}
gene_symbols2 <- f2[, c(1, 11)]
expr2 <- as.data.frame(exp2)
expr2[, 'ID'] <- rownames(expr2)
expr2 <- merge(x = expr2, y = gene_symbols2, by.x = 'ID')
```

#### Adding Index

```{r}
rownames(phe2) <- paste(rownames(phe2), '2', sep='_')
colnames(expr2) <- paste(colnames(expr2), '2', sep='_')
```

#### Removing Duplicated Gene Symbols - Keeping the Mean Gene Expression

```{r}
expre2 <- limma::avereps(expr2[, -c(1, ncol(expr2))], ID = expr2$`Gene Symbol`)
expression2 <- as.data.frame(expre2)
```

#### Filtering of Outcome Groups

```{r}
#ABMR_Status is set to 'ABMR', otherwise set to 'ABMR Stable'
pheno2 <- phe2 %>% mutate(
  ABMR_Status = ifelse(`abmr (no:ch1` == '0_Yes:1): 1', 'ABMR', 'ABMR Stable'),
  #TCMR_Status is set to 'TCMR', otherwise set to 'TCMR Stable'
  TCMR_Status = ifelse(`tcmr (no:ch1` == '0_borderline:1_TCMR:2): 1', 'TCMR', 'TCMR Stable')
  ) %>%
    #Filter out rows with both ABMR and TCMR status
    filter(!(ABMR_Status == 'ABMR' & TCMR_Status == 'TCMR')) %>%
  mutate(
    Outcome = case_when(
      ABMR_Status == 'ABMR Stable' & TCMR_Status == 'TCMR Stable' ~ 'Stable',
      ABMR_Status == 'ABMR Stable' ~ 'TCMR',
      TCMR_Status == 'TCMR Stable' ~ 'ABMR',
      TRUE ~ NA_character_
    )
  )

expression2 <- as.data.frame(expression2)
expression2 <- expression2[, colnames(expression2) %in% rownames(pheno2)]
expression2 <- as.matrix(expression2)
```

### 3. GSE192444

```{r}
#loading dataset3
gse3 <- 'GSE192444'
eSet3 <- getGEO('GSE192444')
eSet3 <- eSet3$GSE192444_series_matrix.txt.gz
exp3 <- exprs(eSet3)
phe3 <- pData(eSet3)
feature3 <- fData(eSet3)
```

#### Cleaning Feature Data

```{r}
f3 <- feature3 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f3$'Gene Symbol' <- sapply(strsplit(f3$'Gene Symbol', ' /// '), function(x) x[1])
```

#### Substituting Gene Probe ID with Gene Symbol

```{r}
gene_symbols3 <- f3[, c(1, 15)]
expr3 <- as.data.frame(exp3)
expr3[, 'ID'] <- rownames(expr3)
expr3 <- merge(x = expr3, y = gene_symbols3, by.x = 'ID')
```

#### Adding Index

```{r}
rownames(phe3) <- paste(rownames(phe3), '3', sep='_')
colnames(expr3) <- paste(colnames(expr3), '3', sep='_')
```

#### Removing Duplicated Gene Symbols - Keeping the Mean Gene Expression

```{r}
expre3 <- limma::avereps(expr3[, -c(1, ncol(expr3))], ID = expr3$`Gene Symbol`)
expression3 <- as.data.frame(expre3)
```

#### Filtering of Outcome Groups

```{r}
# Check if 'mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1' contains 'ABMR'
phe3$Outcome <- ifelse(str_detect(phe3$`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1`, 'ABMR'), 'ABMR',
                       ifelse(str_detect(phe3$`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1`, 'TCMR'), 'TCMR',
                              ifelse(str_detect(phe3$`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1`, 'NR'), 'Stable', 'Other')))
# Filter out rows where 'Outcome' is 'Other'
pheno3 <- phe3 %>% filter(Outcome != 'Other')

expression3 <- as.data.frame(expression3)
expression3 <- expression3[, colnames(expression3) %in% rownames(pheno3)]
expression3 <- as.matrix(expression3)
```

### 4. GSE124203

```{r}
#loading dataset4
gse4 <- 'GSE124203'
eSet4 <- getGEO('GSE124203')
eSet4 <- eSet4$GSE124203_series_matrix.txt.gz
exp4 <- exprs(eSet4)
phe4 <- pData(eSet4)
feature4 <- fData(eSet4)
```

#### Cleaning Feature Data

```{r}
f4 <- feature4 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f4$'Gene Symbol' <- sapply(strsplit(f4$'Gene Symbol', ' /// '), function(x) x[1])
```

#### Substituting Gene Probe ID with Gene Symbol

```{r}
gene_symbols4 <- f4[, c(1, 15)]
expr4 <- as.data.frame(exp4)
expr4[, 'ID'] <- rownames(expr4)
expr4 <- merge(x = expr4, y = gene_symbols4, by.x = 'ID')
```

#### Adding Index

```{r}
rownames(phe4) <- paste(rownames(phe4), '4', sep='_')
colnames(expr4) <- paste(colnames(expr4), '4', sep='_')
```

#### Removing Duplicated Gene Symbols - Keeping the Mean Gene Expression

```{r}
expre4 <- limma::avereps(expr4[, -c(1, ncol(expr4))], ID = expr4$`Gene Symbol`)
expression4 <- as.data.frame(expre4)
```

#### Filtering of Outcome Groups

```{r}
# Check if the specified column contains 'ABMR'
phe4$Outcome <- ifelse(str_detect(phe4$`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1`, 'ABMR'), 'ABMR',
                       ifelse(str_detect(phe4$`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1`, 'TCMR'), 'TCMR',
                              ifelse(str_detect(phe4$`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1`, 'NR'), 'Stable', 'Other')))
# Filtering out rows from 'phe4' where 'Outcome' is 'Other'
pheno4 <- phe4 %>% filter(Outcome != 'Other')

expression4 <- as.data.frame(expression4)
expression4 <- expression4[, colnames(expression4) %in% rownames(pheno4)]
expression4 <- as.matrix(expression4)
```

### 5. GSE98320

```{r}
#loading dataset5
gse5 <- 'GSE98320'
eSet5 <- getGEO('GSE98320')
eSet5 <- eSet5$GSE98320_series_matrix.txt.gz
exp5 <- exprs(eSet5)
phe5 <- pData(eSet5)
feature5 <- fData(eSet5)
```

#### Cleaning Feature Data

```{r}
f5 <- feature5 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f5$'Gene Symbol' <- sapply(strsplit(f5$'Gene Symbol', ' /// '), function(x) x[1])
```

#### Substituting Gene Probe ID with Gene Symbol

```{r}
gene_symbols5 <- f5[, c(1, 17)]
expr5 <- as.data.frame(exp5)
expr5[, 'ID'] <- rownames(expr5)
expr5 <- merge(x = expr5, y = gene_symbols5, by.x = 'ID')
```

#### Adding Index

```{r}
rownames(phe5) <- paste(rownames(phe5), '5', sep='_')
colnames(expr5) <- paste(colnames(expr5), '5', sep='_')
```

#### Removing Duplicated Gene Symbols - Keeping the Mean Gene Expression

```{r}
expre5 <- limma::avereps(expr5[,-c(1, ncol(expr5))], ID = expr5$`Gene Symbol`)
expression5 <- as.data.frame(expre5)
```

#### Filtering of Outcome Groups

```{r}
pheno5 <- phe5 %>% 
  filter(`mmdx:ch1` != '-', `mmdx:ch1` != 'Mixed')%>% 
  # Create a new 'Outcome' column based on the values of 'mmdx:ch1'
  mutate(Outcome = case_when(
  `mmdx:ch1` == 'ABMR' ~ 'ABMR',
  `mmdx:ch1` == 'TCMR' ~ 'TCMR',
  `mmdx:ch1` == 'NR' ~ 'Stable',
  TRUE ~ NA_character_
))

expression5 <- as.data.frame(expression5)
expression5 <- expression5[, colnames(expression5) %in% rownames(pheno5)]
expression5 <- as.matrix(expression5)
```

## Combination

### Selecting the Top 75 Percentile Gene Expressions

```{r}
set.seed(3888)
#Selecting the Top 75 Percentile Gene Expressions
top_75_percentile <- function(eMat) {
  eMat <- as.data.frame(eMat)
  threshold <- quantile(rowSums(eMat), 0.25)
  filtered_eMat <- eMat[rowSums(eMat) > threshold,]
  return (filtered_eMat)
}
Expression1 <- top_75_percentile(expression1)
Expression2 <- top_75_percentile(expression2)
Expression3 <- top_75_percentile(expression3)
Expression4 <- top_75_percentile(expression4)
Expression5 <- top_75_percentile(expression5)
pheno1 <- pheno1[colnames(Expression1),]
pheno2 <- pheno2[colnames(Expression2),]
pheno3 <- pheno3[colnames(Expression3),]
pheno4 <- pheno4[colnames(Expression4),]
pheno5 <- pheno5[colnames(Expression5),]
```

#### Generating an UpSet plot

```{r}
#Upset plot
list_of_expressions <- list(
  'GSE36059' = Expression1,
  'GSE129166' = Expression2,
  'GSE192444' = Expression3,
  'GSE124203' = Expression4,
  'GSE98320' = Expression5
)
binary_transform <- function(eMat) {
  eMat <- !is.na(eMat) & eMat > 0
  return(eMat)
}
binary_data <- lapply(list_of_expressions, binary_transform)
all_genes <- unique(unlist(lapply(binary_data, rownames)))
binary_matrix <- matrix(0, nrow = length(all_genes), ncol = length(binary_data),
                        dimnames = list(all_genes, names(binary_data)))
for (i in seq_along(binary_data)) {
  binary_matrix[rownames(binary_data[[i]]), i] <- 1
}
upset_binary_genes <- as.data.frame(binary_matrix)
colors <- colorRampPalette(brewer.pal(9,'Set1'))(27)
query_list <- list()
tmp <- unique(upset_binary_genes[,c('GSE36059','GSE129166','GSE192444','GSE124203','GSE98320')])
tmp <- tmp[rowSums(tmp)!=0,]
for (i in 1:27) {
  query_list[[i]]<- list(query=intersects,
                         params=list(colnames(tmp)[which(tmp[i,]==1)]),
                         color =colors[i],
                         active=T)
}
upset_plot_before_combine <-upset(upset_binary_genes,
                                  nsets=5,
                                  nintersects=50,
                                  mb.ratio = c(0.6,0.4),
                                  order.by=c('freq'),
                                  decreasing=c(TRUE,FALSE),
                                  sets.bar.color=c('#1879B6',
                                                   '#A5D0E4',
                                                   '#32A34A',
                                                   '#B6D98C',
                                                   '#F5999A'),
                                  shade.color=NA)
upset_plot_before_combine
```

### Combining the Different Datasets

```{r,message=F}
#intersection
common_genes <- Reduce(intersect, list(rownames(Expression1), rownames(Expression2), 
                                       rownames(Expression3), rownames(Expression4), 
                                       rownames(Expression5)))
expression1_common <- expression1[common_genes,]
expression2_common <- expression2[common_genes,]
expression3_common <- expression3[common_genes,]
expression4_common <- expression4[common_genes,]
expression5_common <- expression5[common_genes,]
combined_expression <- cbind(expression1_common, expression2_common, 
                             expression3_common, expression4_common, 
                             expression5_common)
```

### Adding Batch Labels

```{r}
#Add batch labels
batch_labels <- list(
  Expression1 = colnames(expression1_common),
  Expression2 = colnames(expression2_common),
  Expression3 = colnames(expression3_common),
  Expression4 = colnames(expression4_common),
  Expression5 = colnames(expression5_common)
)
names(batch_labels) <- c("1", "2", "3", "4", "5")
```

### Combined Outcomes

```{r}
#combined five datasets outcomes
pheno1 <- pheno1[colnames(expression1_common),]
pheno2 <- pheno2[colnames(expression2_common),]
pheno3 <- pheno3[colnames(expression3_common),]
pheno4 <- pheno4[colnames(expression4_common),]
pheno5 <- pheno5[colnames(expression5_common),]
group1 <- data.frame(Outcome = pheno1$Outcome, row.names = rownames(pheno1))
group2 <- data.frame(Outcome = pheno2$Outcome, row.names = rownames(pheno2))
group3 <- data.frame(Outcome = pheno3$Outcome, row.names = rownames(pheno3))
group4 <- data.frame(Outcome = pheno4$Outcome, row.names = rownames(pheno4))
group5 <- data.frame(Outcome = pheno5$Outcome, row.names = rownames(pheno5))
group <- dplyr::bind_rows(group1, group2, group3, group4, group5)
table(group)
```

#### Generating a Stacked Bar Plot of Samples

```{r}
stack <- data.frame(
  Dataset = rep(c('GSE36059', 'GSE129166', 'GSE192444', 'GSE124203', 'GSE98320'), each=3),
  Outcome = rep(c('ABMR', 'TCMR', 'Stable'), times=5),
  Count = c(table(pheno1$Outcome), table(pheno2$Outcome), 
            table(pheno3$Outcome), table(pheno4$Outcome), table(pheno5$Outcome))
)
stack <- stack %>%
  group_by(Dataset) %>%
  mutate(Proportion = Count / sum(Count)) %>%
  ungroup()
ggplot(stack, aes(x = Dataset, y = Count, fill = Outcome)) +
  geom_bar(stat = 'identity', position = 'stack') +
  labs(title = "Proportion of Outcome Types by Dataset",
       x = "Dataset",
       y = "Proportion of Samples",
       fill = "Outcome Type") +
  theme_minimal()+
  scale_fill_manual(values = c('ABMR' = 'navyblue', 'TCMR' = 'lightblue', "Stable" = 'grey'))
```

### Downsampling to Balance Outcome Groups and Normalising Expression Intensities

```{r}
# Aim: ABMR = TCMR = Stable = 241
set.seed(3888)
ABMR_samples <- sample(colnames(combined_expression)[group == 'ABMR'], 241)
set.seed(3888)
TCMR_samples <- sample(colnames(combined_expression)[group == 'TCMR'], 241)
set.seed(3888)
Stable_samples <- sample(colnames(combined_expression)[group == 'Stable'], 241)

# Combine all selected samples into a single vector
selected_samples <- c(ABMR_samples, TCMR_samples, Stable_samples)
Combined_Expression <- log2(combined_expression[, selected_samples]+1)

# Function to filter batch labels based on the selected samples
filter_batches <- function(batches, samples) {
  lapply(batches, function(batch) batch[batch %in% samples])
}
batch_labels <- filter_batches(batch_labels, selected_samples)

# Subset the 'group' data frame to include only the selected samples
group <- group[selected_samples, , drop = FALSE]
batch_info <- apply(sapply(batch_labels, FUN = `%in%`, x = rownames(group)), MARGIN = 1, FUN = which)
```

#### Generating a Stacked Bar Plot after Downsampling

```{r}
group_balance <- group %>% filter(rownames(group) %in% colnames(Combined_Expression))
Sample <- rownames(group_balance)
group_balance <- group_balance %>%
  mutate(Dataset = case_when(
    grepl("_1", Sample) ~ "GSE36059",
    grepl("_2", Sample) ~ "GSE129166",
    grepl("_3", Sample) ~ "GSE192444",
    grepl("_4", Sample) ~ "GSE124203",
    grepl("_5", Sample) ~ "GSE98320"
  ))
ggplot(group_balance, aes(x = Dataset, fill = Outcome)) +
  geom_bar(position = 'stack') +
  labs(title = 'Proportion of Outcome Types by Dataset after Downsampling', 
       x = 'Dataset', 
       y = 'Proportion of Samples') +
  theme_minimal() +
  scale_fill_manual(values = c('ABMR' = 'navyblue', 'Stable' = 'grey', 'TCMR' = 'lightblue'))
```

### Splitting Original Data - 70% for Training, 30% for Testing

```{r}
#randomly selecting 70%
set.seed(3888)
n <- ncol(Combined_Expression)
idx <- sample(n)
set.seed(3888)
split_idx<- floor(0.3*n)
eMat_30 <-Combined_Expression[, idx[1:split_idx]]
eMat_70 <- Combined_Expression[, idx[(split_idx + 1):n]]
group_70 <- group %>% filter(rownames(group) %in% colnames(eMat_70))
table(group_70)
```

#### Generating Stacked Bar Plot after Splitting

```{r}
Sample_split <- rownames(group_70)
group_70 <- group_70 %>%
  mutate(Dataset = case_when(
    grepl("_1", Sample_split) ~ "GSE36059",
    grepl("_2", Sample_split) ~ "GSE129166",
    grepl("_3", Sample_split) ~ "GSE192444",
    grepl("_4", Sample_split) ~ "GSE124203",
    grepl("_5", Sample_split) ~ "GSE98320"
  ))
ggplot(group_70, aes(x = Dataset,fill = Outcome)) +
  geom_bar(position = 'stack') +
  labs(title = "Proportion of Outcome Types by Dataset after Splitting",
       x = "Dataset",
       y = "Proportion of Samples",
       fill = "Outcome Type") +
  theme_minimal()+
  scale_fill_manual(values = c('ABMR' = 'navyblue', 'TCMR' = 'lightblue', "Stable" = 'grey'))
table(group %>% filter(rownames(group) %in% colnames(eMat_70)))
```

### Adjusting the Batch Labels

```{r}
#Adjust batch labels
batch_info_70 <- apply(sapply(batch_labels, FUN = `%in%`, x = colnames(eMat_70)), MARGIN = 1, FUN = which)
pca_batch_info = case_when(
  batch_info_70 == 1 ~ "GSE36059",
  batch_info_70 == 2 ~ "GSE129166",
  batch_info_70 == 3 ~ "GSE192444",
  batch_info_70 == 4 ~ "GSE124203",
  batch_info_70 == 5 ~ "GSE98320"
)
```

#### Exploring Data with PCA - Before Batch Correction

```{r}
set.seed(3888)
gse_pca <- prcomp(t(eMat_70))
df_toplot_before <- data.frame(Outcome = group[colnames(eMat_70),],
                        pc1 = gse_pca$x[,1], pc2 = gse_pca$x[,2],
                        batch = pca_batch_info)
#coloured by Outcome
ggplot(df_toplot_before, aes(x = pc1, y = pc2, color = Outcome)) + 
  geom_point(size = 1.5) + labs(colour = "Outcome") +
  theme_minimal() +
  xlab("PC1") + ylab("PC2") + ggtitle("PCA plot - Batch - Outcome")

#coloured by dats set
ggplot(df_toplot_before, aes(x = pc1, y = pc2, color = factor(batch))) + 
  geom_point(size = 1.5) + labs(colour = "Batch") +
  theme_minimal() +
  xlab("PC1") + ylab("PC2") + ggtitle("PCA plot - Batch - Dataset")
```

### ComBat Batch Correction

```{r, echo=FALSE,message=FALSE,warning=FALSE}
#Combat batch correction
set.seed(3888)
MOD <- model.matrix(~factor(group[colnames(eMat_70),]))
combat_eMat_70 <- suppressMessages(
  sva::ComBat(dat = eMat_70, batch = batch_info_70, mod = MOD, par.prior = TRUE, prior.plots = FALSE)
)
dim(combat_eMat_70)
#11837 genes and 507 samples in the final training set
```

#### Exploring Data with PCA - after Batch Correction

```{r}
set.seed(3888)
gse_pca <- prcomp(t(combat_eMat_70))
df_toplot <- data.frame(Outcome = group[colnames(combat_eMat_70),],
                        pc1 = gse_pca$x[,1], pc2 = gse_pca$x[,2],
                        batch = pca_batch_info)
ggplot(df_toplot, aes(x = pc1, y = pc2, color = Outcome)) + 
  geom_point(size = 1.5) + labs(colour = "Outcome") +
  geom_hline(yintercept = 0, linetype = 'dashed',colour='gray') +
  geom_vline(xintercept = 0, linetype = 'dashed',colour='gray')+
  theme_minimal() +
  stat_ellipse(aes(fill = Outcome), geom = 'polygon', alpha = 0.05) +
  xlab("PC1") + ylab("PC2") + ggtitle("PCA plot - Batch - Outcome")

ggplot(df_toplot, aes(x = pc1, y = pc2, color = factor(batch))) + 
  geom_point(size = 1.5) + labs(colour = "Batch") +
  geom_hline(yintercept = 0, linetype = 'dashed',colour='gray') +
  geom_vline(xintercept = 0, linetype = 'dashed',colour='gray')+
  theme_minimal() +
  stat_ellipse( geom = 'polygon', alpha = 0.02) +
  xlab("PC1") + ylab("PC2") + ggtitle("PCA plot - Batch - Dataset")
```

## Model Evaluation with Cross-Validation

### Function to Select Top DE Genes Using Limma

```{r}
set.seed(3888)

# Define a function to identify top genes using differential expression analysis
limma_top_genes <- function(eMat, outcome, n) {
  group_factor = factor(outcome)
  # Create a design matrix for the linear model, omitting the intercept
  design_matrix <- model.matrix(~ 0 + group_factor)
  colnames(design_matrix) <- levels(group_factor)
   # the mean differences between groups   
  contrast_matrix <- makeContrasts(
    ABMR_vs_Stable_and_TCMR = ABMR - (1 / 2 * (Stable + TCMR)),
    TCMR_vs_Stable_and_ABMR = TCMR - (1 / 2 * (Stable + ABMR)),
    Stable_vs_Rejection = Stable - (1 / 2 * (ABMR + TCMR)),
    levels = colnames(design_matrix)
  )
  # Fit the linear model to the expression matrix 'eMat'    
  fit <- lmFit(eMat, design = design_matrix)
  fit <- contrasts.fit(fit, contrasts = contrast_matrix)
  fit <- eBayes(fit)
  
  # Extract the top 'n' genes for each contrast based on the eBayes results
  tT_ABMR <- topTable(fit, n = n, coef = 1)
  tT_TCMR <- topTable(fit, n = n, coef = 2)
  tT_Stable <- topTable(fit, n = n, coef = 3)
  top_genes <- unique(c(rownames(tT_ABMR), rownames(tT_TCMR), rownames(tT_Stable)))
  
  return (top_genes)
}
```

### Function to Evaluate Classifier Performance

```{r}
evaluatePerformance <- function(confusionMatrix) {
  TP_ABMR = confusionMatrix[1, 1]
  FP_ABMR = confusionMatrix[2, 1] + confusionMatrix[3, 1]
  FN_ABMR = confusionMatrix[1, 2] + confusionMatrix[1, 3]
  TN_ABMR = confusionMatrix[2, 2] + confusionMatrix[2, 3] + confusionMatrix[3, 2] + confusionMatrix[3, 3]
  
  precision_ABMR = TP_ABMR / (TP_ABMR + FP_ABMR)
  recall_ABMR = TP_ABMR / (TP_ABMR + FN_ABMR)
  specificity_ABMR = TN_ABMR / (TN_ABMR + FP_ABMR)
  
  TP_Stable = confusionMatrix[2, 2]
  FP_Stable = confusionMatrix[1, 2] + confusionMatrix[3, 2]
  FN_Stable = confusionMatrix[2, 1] + confusionMatrix[2, 3]
  TN_Stable = confusionMatrix[1, 1] + confusionMatrix[1, 3] + confusionMatrix[3, 1] + confusionMatrix[3, 3]
  
  precision_Stable = TP_Stable / (TP_Stable + FP_Stable)
  recall_Stable = TP_Stable / (TP_Stable + FN_Stable)
  specificity_Stable = TN_Stable / (TN_Stable + FP_Stable)
  
  TP_TCMR = confusionMatrix[3, 3]
  FP_TCMR = confusionMatrix[1, 3] + confusionMatrix[2, 3]
  FN_TCMR = confusionMatrix[3, 1] + confusionMatrix[3, 2]
  TN_TCMR = confusionMatrix[1, 1] + confusionMatrix[1, 2] + confusionMatrix[2, 1] + confusionMatrix[2, 2]
  
  precision_TCMR = TP_TCMR / (TP_TCMR + FP_TCMR)
  recall_TCMR = TP_TCMR / (TP_TCMR + FN_TCMR)
  specificity_TCMR = TN_TCMR / (TN_TCMR + FP_TCMR)
  
  accuracy = sum(diag(confusionMatrix)) / sum(confusionMatrix)
  precision = (precision_ABMR + precision_Stable + precision_TCMR) / 3
  recall = (recall_ABMR + recall_Stable + recall_TCMR) / 3
  specificity = (specificity_ABMR + specificity_Stable + specificity_TCMR) / 3
  f1 = (2 * precision * recall) / (precision + recall)
  
  return (c(accuracy, precision, recall, specificity, f1))
}
```

### Repeated 5-fold Cross-Validation

```{r, cache=TRUE}
set.seed(3888)
X = t(eMat_70)
y = factor(group[colnames(eMat_70),])
cvK = 5
n_sim = 20
# Calculate mean accuracy for each model and feature size across all folds
cv_accuracy30_rf = cv_accuracy50_rf = cv_accuracy80_rf = cv_accuracy110_rf = list()
cv_accuracy30_svm = cv_accuracy50_svm = cv_accuracy80_svm = cv_accuracy110_svm = list()
cv_accuracy30_knn2 = cv_accuracy50_knn2 = cv_accuracy80_knn2 = cv_accuracy110_knn2 = list()
cv_accuracy30_knn5 = cv_accuracy50_knn5 = cv_accuracy80_knn5 = cv_accuracy110_knn5 = list()
for (i in 1:n_sim) {
  cvSets = cvFolds(nrow(X), cvK)
  
  cv_accuracy_folds30_rf = cv_accuracy_folds50_rf = cv_accuracy_folds80_rf = cv_accuracy_folds110_rf = list()
  cv_accuracy_folds30_svm = cv_accuracy_folds50_svm = cv_accuracy_folds80_svm = cv_accuracy_folds110_svm = list()
  cv_accuracy_folds30_knn2 = cv_accuracy_folds50_knn2 = cv_accuracy_folds80_knn2 = cv_accuracy_folds110_knn2 = list()
  cv_accuracy_folds30_knn5 = cv_accuracy_folds50_knn5 = cv_accuracy_folds80_knn5 = cv_accuracy_folds110_knn5 = list()
  
  for (j in 1:cvK) {
    test_id = cvSets$subsets[cvSets$which == j]
    X_train = X[-test_id,]
    X_test = X[test_id,]
    y_train = y[-test_id]
    y_test = y[test_id]
    
    set.seed(3888)
    # Applying batch correction to the training data
    batch_train <- apply(sapply(batch_labels, FUN = `%in%`, x = rownames(X_train)), MARGIN = 1, FUN = which)
    mod <- model.matrix(~factor(group[rownames(X_train),]))
    X_train <- suppressMessages(t(sva::ComBat(dat = t(X_train), batch = batch_train, mod = mod, par.prior = TRUE, prior.plots = FALSE)))
    
    # Feature selection with Limma
    DE_genes30 <- limma_top_genes(t(X_train), y_train, 10)
    DE_genes50 <- limma_top_genes(t(X_train), y_train, 17)
    DE_genes80 <- limma_top_genes(t(X_train), y_train, 27)
    DE_genes110 <- limma_top_genes(t(X_train), y_train, 37)
    
    X_train30 = X_train[,DE_genes30]
    X_test30 = X_test[,DE_genes30]
    
    X_train50 = X_train[,DE_genes50]
    X_test50 = X_test[,DE_genes50]
    
    X_train80 = X_train[,DE_genes80]
    X_test80 = X_test[,DE_genes80]
    
    X_train110 = X_train[,DE_genes110]
    X_test110 = X_test[,DE_genes110]
    
    # Random Forest
    rf_fit30 = randomForest(x = X_train30, y = y_train)
    predictions_rf30 = predict(rf_fit30, X_test30)
    cv_accuracy_folds30_rf[[j]] = evaluatePerformance(
      table(Actual = y_test, Prediction = predictions_rf30))
    
    rf_fit50 = randomForest(x = X_train50, y = y_train)
    predictions_rf50 = predict(rf_fit50, X_test50)
    cv_accuracy_folds50_rf[[j]] = evaluatePerformance(
      table(Actual = y_test, Prediction = predictions_rf50))
    
    rf_fit80 = randomForest(x = X_train80, y = y_train)
    predictions_rf80 = predict(rf_fit80, X_test80)
    cv_accuracy_folds80_rf[[j]] = evaluatePerformance(
      table(Actual = y_test, Prediction = predictions_rf80))
    
    rf_fit110 = randomForest(x = X_train110, y = y_train)
    predictions_rf110 = predict(rf_fit110, X_test110)
    cv_accuracy_folds110_rf[[j]] = evaluatePerformance(
      table(Actual = y_test, Prediction = predictions_rf110))
    
    # SVM
    svm_fit30 = svm(X_train30, y_train)
    predictions_svm30 = predict(svm_fit30, X_test30)
    cv_accuracy_folds30_svm[[j]] = evaluatePerformance(
      table(Actual = y_test, Prediction = predictions_svm30))
    
    svm_fit50 = svm(X_train50, y_train)
    predictions_svm50 = predict(svm_fit50, X_test50)
    cv_accuracy_folds50_svm[[j]] = evaluatePerformance(
      table(Actual = y_test, Prediction = predictions_svm50))
    
    svm_fit80 = svm(X_train80, y_train)
    predictions_svm80 = predict(svm_fit80, X_test80)
    cv_accuracy_folds80_svm[[j]] = evaluatePerformance(
      table(Actual = y_test, Prediction = predictions_svm80))
    
    svm_fit110 = svm(X_train110, y_train)
    predictions_svm110 = predict(svm_fit110, X_test110)
    cv_accuracy_folds110_svm[[j]] = evaluatePerformance(
      table(Actual = y_test, Prediction = predictions_svm110))
    
    # KNN (k = 2)
    knn2_fit30 = class::knn(train = X_train30, test = X_test30, cl = y_train, k = 2)
    cv_accuracy_folds30_knn2[[j]] = evaluatePerformance(
      table(Actual = y_test, Prediction = knn2_fit30))
    
    knn2_fit50 = class::knn(train = X_train50, test = X_test50, cl = y_train, k = 2)
    cv_accuracy_folds50_knn2[[j]] = evaluatePerformance(
      table(Actual = y_test, Prediction = knn2_fit50))
    
    knn2_fit80 = class::knn(train = X_train80, test = X_test80, cl = y_train, k = 2)
    cv_accuracy_folds80_knn2[[j]] = evaluatePerformance(
      table(Actual = y_test, Prediction = knn2_fit80))
    
    knn2_fit110 = class::knn(train = X_train110, test = X_test110, cl = y_train, k = 2)
    cv_accuracy_folds110_knn2[[j]] = evaluatePerformance(
      table(Actual = y_test, Prediction = knn2_fit110))
    
    # KNN (k = 5)
    knn5_fit30 = class::knn(train = X_train30, test = X_test30, cl = y_train, k = 5)
    cv_accuracy_folds30_knn5[[j]] = evaluatePerformance(
      table(Actual = y_test, Prediction = knn5_fit30))
    
    knn5_fit50 = class::knn(train = X_train50, test = X_test50, cl = y_train, k = 5)
    cv_accuracy_folds50_knn5[[j]] = evaluatePerformance(
      table(Actual = y_test, Prediction = knn5_fit50))
    
    knn5_fit80 = class::knn(train = X_train80, test = X_test80, cl = y_train, k = 5)
    cv_accuracy_folds80_knn5[[j]] = evaluatePerformance(
      table(Actual = y_test, Prediction = knn5_fit80))
    
    knn5_fit110 = class::knn(train = X_train110, test = X_test110, cl = y_train, k = 5)
    cv_accuracy_folds110_knn5[[j]] = evaluatePerformance(
      table(Actual = y_test, Prediction = knn5_fit110))
  }
  
  cv_accuracy30_rf[[i]] = colMeans(do.call(rbind, cv_accuracy_folds30_rf))
  cv_accuracy50_rf[[i]] = colMeans(do.call(rbind, cv_accuracy_folds50_rf))
  cv_accuracy80_rf[[i]] = colMeans(do.call(rbind, cv_accuracy_folds80_rf))
  cv_accuracy110_rf[[i]] = colMeans(do.call(rbind, cv_accuracy_folds110_rf))
  
  cv_accuracy30_svm[[i]] = colMeans(do.call(rbind, cv_accuracy_folds30_svm))
  cv_accuracy50_svm[[i]] = colMeans(do.call(rbind, cv_accuracy_folds50_svm))
  cv_accuracy80_svm[[i]] = colMeans(do.call(rbind, cv_accuracy_folds80_svm))
  cv_accuracy110_svm[[i]] = colMeans(do.call(rbind, cv_accuracy_folds110_svm))
  
  cv_accuracy30_knn2[[i]] = colMeans(do.call(rbind, cv_accuracy_folds30_knn2))
  cv_accuracy50_knn2[[i]] = colMeans(do.call(rbind, cv_accuracy_folds50_knn2))
  cv_accuracy80_knn2[[i]] = colMeans(do.call(rbind, cv_accuracy_folds80_knn2))
  cv_accuracy110_knn2[[i]] = colMeans(do.call(rbind, cv_accuracy_folds110_knn2))
  cv_accuracy30_knn5[[i]] = colMeans(do.call(rbind, cv_accuracy_folds30_knn5))
  cv_accuracy50_knn5[[i]] = colMeans(do.call(rbind, cv_accuracy_folds50_knn5))
  cv_accuracy80_knn5[[i]] = colMeans(do.call(rbind, cv_accuracy_folds80_knn5))
  cv_accuracy110_knn5[[i]] = colMeans(do.call(rbind, cv_accuracy_folds110_knn5))
}
cv_accuracy30_rf <- do.call(rbind, cv_accuracy30_rf)
cv_accuracy50_rf <- do.call(rbind, cv_accuracy50_rf)
cv_accuracy80_rf <- do.call(rbind, cv_accuracy80_rf)
cv_accuracy110_rf <- do.call(rbind, cv_accuracy110_rf)
cv_accuracy30_svm <- do.call(rbind, cv_accuracy30_svm)
cv_accuracy50_svm <- do.call(rbind, cv_accuracy50_svm)
cv_accuracy80_svm <- do.call(rbind, cv_accuracy80_svm)
cv_accuracy110_svm <- do.call(rbind, cv_accuracy110_svm)
cv_accuracy30_knn2 <- do.call(rbind, cv_accuracy30_knn2)
cv_accuracy50_knn2 <- do.call(rbind, cv_accuracy50_knn2)
cv_accuracy80_knn2 <- do.call(rbind, cv_accuracy80_knn2)
cv_accuracy110_knn2 <- do.call(rbind, cv_accuracy110_knn2)
cv_accuracy30_knn5 <- do.call(rbind, cv_accuracy30_knn5)
cv_accuracy50_knn5 <- do.call(rbind, cv_accuracy50_knn5)
cv_accuracy80_knn5 <- do.call(rbind, cv_accuracy80_knn5)
cv_accuracy110_knn5 <- do.call(rbind, cv_accuracy110_knn5)
```

### Visualization Evaluation

```{r, cache=TRUE}
combine_data <- function(data, method, size) {
  data %>%
    as.data.frame() %>%
    mutate(Method = method, Size = size)
}

# Create a list of data frames
df_list <- list(
  combine_data(cv_accuracy30_rf, 'RF', 30),
  combine_data(cv_accuracy50_rf, 'RF', 50),
  combine_data(cv_accuracy80_rf, 'RF', 80),
  combine_data(cv_accuracy110_rf, 'RF', 110),
  combine_data(cv_accuracy30_svm, 'SVM', 30),
  combine_data(cv_accuracy50_svm, 'SVM', 50),
  combine_data(cv_accuracy80_svm, 'SVM', 80),
  combine_data(cv_accuracy110_svm, 'SVM', 110),
  combine_data(cv_accuracy30_knn2, 'KNN2', 30),
  combine_data(cv_accuracy50_knn2, 'KNN2', 50),
  combine_data(cv_accuracy80_knn2, 'KNN2', 80),
  combine_data(cv_accuracy110_knn2, 'KNN2', 110),
  combine_data(cv_accuracy30_knn5, 'KNN5', 30),
  combine_data(cv_accuracy50_knn5, 'KNN5', 50),
  combine_data(cv_accuracy80_knn5, 'KNN5', 80),
  combine_data(cv_accuracy110_knn5, 'KNN5', 110)
)
all_data <- bind_rows(df_list) %>%
  pivot_longer(cols = -c(Method, Size), names_to = 'Metric', values_to = 'Value')
create_boxplot <- function(metric_name, y_label, plot_title) {
  ggplot(all_data %>% filter(Metric == metric_name), aes(x = interaction(Method, Size), y = Value, fill = Method)) +
    geom_boxplot() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
    labs(x = NULL, y = y_label, title = plot_title) +
    scale_fill_manual(values = c('RF' = '#1879B6', 'SVM' = '#A5D0E4', 'KNN2' = '#32A34A', 'KNN5' = '#B6D98C'))
}
p1 <- create_boxplot('V1', 'Accuracy', 'Distribution of Mean 5-fold CV Accuracies')
p2 <- create_boxplot('V2', 'Precision', 'Distribution of Mean 5-fold CV Precisions')
p3 <- create_boxplot('V3', 'Sensitivity', 'Distribution of Mean 5-fold CV Sensitivities')
p4 <- create_boxplot('V4', 'Specificity', 'Distribution of Mean 5-fold CV Specificities')
p5 <- create_boxplot('V5', 'F1 score', 'Distribution of Mean 5-fold CV F1 Scores')

print(p1)
print(p2)
print(p3)
print(p4)
print(p5)
```


### Evaluating with Table

```{r, cache=TRUE}
# Create a data frame to compare model performances
comparision <- data.frame(Model = c("RF - 30", "RF - 50", "RF - 80", "RF - 110",
                                    "SVM - 30", "SVM - 50", "SVM - 80", "SVM - 110",
                                    "KNN2 - 30", "KNN2 - 50", "KNN2 - 80", "KNN2 - 110",
                                    "KNN5 - 30", "KNN5 - 50", "KNN5 - 80", "KNN5 - 110"),
                          Mean = c(mean(cv_accuracy30_rf[,1]), mean(cv_accuracy50_rf[,1]),
                                   mean(cv_accuracy80_rf[,1]), mean(cv_accuracy110_rf[,1]),
                                   mean(cv_accuracy30_svm[,1]), mean(cv_accuracy50_svm[,1]),
                                   mean(cv_accuracy80_svm[,1]), mean(cv_accuracy110_svm[,1]),
                                   mean(cv_accuracy30_knn2[,1]), mean(cv_accuracy50_knn2[,1]),
                                   mean(cv_accuracy80_knn2[,1]), mean(cv_accuracy110_knn2[,1]),
                                   mean(cv_accuracy30_knn5[,1]), mean(cv_accuracy50_knn5[,1]),
                                   mean(cv_accuracy80_knn5[,1]), mean(cv_accuracy110_knn5[,1])),
                          Variance = c(var(cv_accuracy30_rf[,1]), var(cv_accuracy50_rf[,1]),
                                       var(cv_accuracy80_rf[,1]), var(cv_accuracy110_rf[,1]),
                                       var(cv_accuracy30_svm[,1]), var(cv_accuracy50_svm[,1]),
                                       var(cv_accuracy80_svm[,1]), var(cv_accuracy110_svm[,1]),
                                       var(cv_accuracy30_knn2[,1]), var(cv_accuracy50_knn2[,1]),
                                       var(cv_accuracy80_knn2[,1]), var(cv_accuracy110_knn2[,1]),
                                       var(cv_accuracy30_knn5[,1]), var(cv_accuracy50_knn5[,1]),
                                       var(cv_accuracy80_knn5[,1]), var(cv_accuracy110_knn5[,1])))
comparision %>% arrange(Variance)
```

## Final model - Random Forest with Top 50 DE Genes

### Selecting the Top 50 DE Genes Ssing Limma

```{r, cache=TRUE}
set.seed(3888)
# Convert the group variable to a factor based on the columns of combat_eMat_70
group_factor_final = factor(group[colnames(combat_eMat_70),])
design_matrix_final <- model.matrix(~ 0 + group_factor_final)
colnames(design_matrix_final) <- levels(group_factor_final)

# Define the contrast matrix to specify the comparisons of interest
contrast_matrix_final <- makeContrasts(
    ABMR_vs_Stable_and_TCMR = ABMR - (1 / 2 * (Stable + TCMR)),
    TCMR_vs_Stable_and_ABMR = TCMR - (1 / 2 * (Stable + ABMR)),
    Stable_vs_Rejection = Stable - (1 / 2 * (ABMR + TCMR)),
    levels = colnames(design_matrix_final)
  )
# Fit the linear model to the batch-corrected expression data
Fit <- lmFit(combat_eMat_70, design = design_matrix_final)
Fit <- contrasts.fit(Fit, contrasts = contrast_matrix_final)
Fit <- eBayes(Fit)
#choosing 50 genes among 3 groups == 17 for each group and we find the union
tT_ABMR_final <- topTable(Fit, n = 17, coef = 1, sort.by='p')
tT_TCMR_final <- topTable(Fit, n = 17, coef = 2, sort.by='p')
tT_Stable_final <- topTable(Fit, n = 17, coef = 3, sort.by='p')
top_genes_final <- unique(c(rownames(tT_ABMR_final), rownames(tT_TCMR_final), rownames(tT_Stable_final)))
length(top_genes_final)# there are 2 genes appearring in 2 groups
top_genes_final
genes49_expression <- combat_eMat_70[top_genes_final,]
```
#### Heatmap for 49 Genes

```{r,message=FALSE, cache=TRUE}
library(ComplexHeatmap)
library(RColorBrewer)
heatmap_colors <- colorRampPalette(brewer.pal(9, "Blues"))(255)
pheatmap(genes49_expression, 
         cluster_rows = TRUE, 
         cluster_cols = TRUE, 
         scale = "row", 
         show_rownames = TRUE, 
         show_colnames = FALSE,
         main = "Heatmap of Top 49 Genes",
         color = heatmap_colors,
         fontsize_row = 6.5,
         heatmap_legend_param = list(title = NULL))
```

### Evaluating the Top 1 Gene

```{r, cache=TRUE}
# Extract the top gene for the ABMR vs. (Stable and TCMR) contrast, sorted by p-value
top1_abmr <- topTable(Fit, n = 1, coef = 1,sort.by='p')
top1_tcmr <- topTable(Fit, n = 1, coef = 2,sort.by='p')
top1_stable <- topTable(Fit, n = 1, coef = 3,sort.by='p')

# Combine the row names (gene names) of the top genes from each contrast
top1_genes_final <- unique(c(rownames(top1_abmr), rownames(top1_tcmr),rownames(top1_stable)))
print('The top 3 union genes are:')
top1_genes_final
# Extract the top gene by sorting based on the F-statistic
total_1 <- rownames(topTable(Fit,sort.by="F",n=1))
print('The top gene by sorting F is:')
total_1
```

### Finding Intersectiion Genes
```{r, cache=TRUE}
# Find the intersection of top genes between ABMR and TCMR
abmr_tcmr <-intersect(rownames(tT_ABMR_final), rownames(tT_TCMR_final))
length(abmr_tcmr)
# Find the intersection of top genes between ABMR and Stable
abmr_stable<- intersect(rownames(tT_ABMR_final),rownames(tT_Stable_final))
length(abmr_stable)
paste('The common gene among ABMR and Stable is:',abmr_stable)
# Find the intersection of top genes between TCMR and Stable
tcmr_stable <- intersect(rownames(tT_TCMR_final),rownames(tT_Stable_final))
length(tcmr_stable)
paste('The common gene among TCMR and Stable is:',tcmr_stable)
```

### Preparing the Training and Testing Data

```{r, cache=TRUE}
set.seed(3888)
X_train_final_model <- t(combat_eMat_70[rownames(combat_eMat_70) %in% top_genes_final, ])
y_train_final_model <- factor(group[rownames(X_train_final_model),])
X_test_final_model <- eMat_30[rownames(eMat_30) %in% top_genes_final, ]
X_test_final_model <- t(X_test_final_model)
y_test_final_model <- factor(group[rownames(X_test_final_model),])
```

### Building the Final Model and Prediction on Unseen Test Data

```{r, cache=TRUE}
set.seed(3888)
rf_fit_final <- randomForest(x = X_train_final_model, y = y_train_final_model)
predicted_final_model <- predict(rf_fit_final, newdata=X_test_final_model)
confusionMatrix(as.factor(predicted_final_model), as.factor(y_test_final_model))
```

### Evaluation of the Final Model

```{r, cache=TRUE}
# Evaluate the performance of the final model
performance_final_model = evaluatePerformance(
  table(Actual = y_test_final_model, Prediction = predicted_final_model)
)
# Create a data frame to store the performance metrics
performance_df = data.frame(
  Metric = c("Accuracy", "Precision", "Sensitivity", "Specificity", "F1 Score"),
  Value = performance_final_model)
# Plot the performance metrics using ggplot2
ggplot(performance_df, aes(x = Metric, y = Value)) +
  geom_bar(stat = "identity", width = 0.4, fill = "#1879B6") +
  geom_text(aes(label = round(Value, 5)), vjust = -0.5) +
  labs(title = "Performance Metrics of Final Model") +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5))
```

### ROC-AUC Curve

```{r, message=FALSE, warning=FALSE, cache=TRUE}
# Recode the actual test labels to binary factors for each class
response_abmr <- factor(ifelse(y_test_final_model == 'ABMR', 'ABMR', 'Not ABMR'))
response_tcmr <- factor(ifelse(y_test_final_model == 'TCMR', 'TCMR', 'Not TCMR'))
response_stable <- factor(ifelse(y_test_final_model == 'Stable', 'Stable', 'Not Stable'))
# Get the probability predictions from the final random forest model
prob_predictions <- predict(rf_fit_final, newdata = X_test_final_model, type = "prob")
roc_abmr <- suppressMessages(roc(response = response_abmr, 
                predictor = prob_predictions[, 'ABMR']))
roc_tcmr <- suppressMessages(roc(response = response_tcmr, 
                predictor = prob_predictions[, 'TCMR']))
roc_stable <- suppressMessages(roc(response = response_stable, 
                  predictor = prob_predictions[, 'Stable']))

# Plot the ROC curve for ABMR
plot(roc_abmr, col = "#F5999A", main = 'ROC Curves for Each Class')
plot(roc_tcmr, col = "#1879B6", add = TRUE)
plot(roc_stable, col = "#32A34A", add = TRUE)
# Add a legend to the plot to identify each ROC curve and display the AUC
legend('bottomright', legend = c(
  paste('ABMR vs. Others:',round(roc_abmr$auc,3)),
  paste('TCMR vs. Others:',round(roc_tcmr$auc,3)),
  paste('Stable vs. Others:',round(roc_stable$auc,3))),
       col = c('#F5999A', '#1879B6', '#32A34A'), 
  lwd = 2,
  cex = 0.8)
grid(col = 'lightgray')
```