---
title: "DATA3888 Interdisciplinary Project"
author: 'Kidney20'
output:
  html_document:
    fig_caption: yes
    self_contained: yes
    theme: flatly
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
---

# Loading required packages

```{r setup, include=FALSE}
library(caret)
library(cvTools)
library(dplyr)
library(e1071)
library(GEOquery)
library(ggplot2)
library(glmnet)
library(limma)
library(pheatmap)
library(randomForest)
library(stringr)
library(sva)
library(tidyverse)
library(pROC)
library(precrec)

rm(list = ls())
```

# Loading datasets and cleaning

## 1. GSE36059

```{r}
gse1 <- 'GSE36059'
eSet1 <- getGEO('GSE36059')
eSet1 <- eSet1$GSE36059_series_matrix.txt.gz
exp1 <- exprs(eSet1)
phe1 <- pData(eSet1)
feature1 <- fData(eSet1)
```

### Cleaning feature data

```{r}
f1 <- feature1 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f1$'Gene Symbol' <- sapply(strsplit(f1$'Gene Symbol', ' /// '), function(x) x[1])
dim(f1)
```

### Substituting gene probe ID with gene symbol

```{r}
gene_symbols1 <- f1[, c(1, 11)]
expr1 <- as.data.frame(exp1)
expr1[, 'ID'] <- rownames(expr1)
expr1 <- merge(x = expr1, y = gene_symbols1, by.x = 'ID')
dim(expr1)
```

### Adding index

```{r}
rownames(phe1) <- paste(rownames(phe1), '1', sep='_')
colnames(expr1) <- paste(colnames(expr1), '1', sep='_')
dim(phe1)
```

### Removing duplicated gene symbols - keeping the mean gene expression

```{r}
expre1 <- limma::avereps(expr1[, -c(1, ncol(expr1))], ID = expr1$`Gene Symbol`)
expression1 <- as.data.frame(expre1)
dim(expression1)
```

### Filtering of outcome groups

```{r}
phe1$Outcome <- ifelse(str_detect(phe1$title, 'non-rejecting'), 'Stable',
                       ifelse(str_detect(phe1$title, 'ABMR'), 'ABMR',
                              ifelse(str_detect(phe1$title, 'TCMR'), 'TCMR', 'Other')))
pheno1 <- phe1 %>% filter(Outcome != 'Other')
table(pheno1$Outcome)

expression1 <- as.data.frame(expression1)
expression1 <- expression1[, colnames(expression1) %in% rownames(pheno1)]
expression1 <- as.matrix(expression1)
```

## 2. GSE129166

```{r}
gse2 <- 'GSE129166'
eSet2 <- getGEO('GSE129166')
eSet2 <- eSet2$GSE129166_series_matrix.txt.gz
exp2 <- exprs(eSet2)
phe2 <- pData(eSet2)
feature2 <- fData(eSet2)
```

### Filtering to biopsy data only

```{r}
phe2 <- phe2 %>% filter(str_detect(title, 'BIOPSY'))
exp2 <- exp2[, rownames(phe2)]
dim(exp2)
```

### Cleaning feature data

```{r}
f2 <- feature2 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f2$'Gene Symbol' <- sapply(strsplit(f2$'Gene Symbol', ' /// '), function(x) x[1])
dim(f2)
```

### Substituting gene probe ID with gene symbol

```{r}
gene_symbols2 <- f2[, c(1, 11)]
expr2 <- as.data.frame(exp2)
expr2[, 'ID'] <- rownames(expr2)
expr2 <- merge(x = expr2, y = gene_symbols2, by.x = 'ID')
dim(expr2)
```

### Adding index

```{r}
rownames(phe2) <- paste(rownames(phe2), '2', sep='_')
colnames(expr2) <- paste(colnames(expr2), '2', sep='_')
dim(phe2)
```

### Removing duplicated gene symbols - keeping the mean gene expression

```{r}
expre2 <- limma::avereps(expr2[, -c(1, ncol(expr2))], ID = expr2$`Gene Symbol`)
expression2 <- as.data.frame(expre2)
dim(expression2)
```

### Filtering of outcome groups

```{r}
pheno2 <- phe2 %>% mutate(
  ABMR_Status = ifelse(`abmr (no:ch1` == '0_Yes:1): 1', 'ABMR', 'ABMR Stable'),
  TCMR_Status = ifelse(`tcmr (no:ch1` == '0_borderline:1_TCMR:2): 1', 'TCMR', 'TCMR Stable')
  ) %>%
    filter(!(ABMR_Status == 'ABMR' & TCMR_Status == 'TCMR')) %>%
  mutate(
    Outcome = case_when(
      ABMR_Status == 'ABMR Stable' & TCMR_Status == 'TCMR Stable' ~ 'Stable',
      ABMR_Status == 'ABMR Stable' ~ 'TCMR',
      TCMR_Status == 'TCMR Stable' ~ 'ABMR',
      TRUE ~ NA_character_
    )
  )
table(pheno2$Outcome)

expression2 <- as.data.frame(expression2)
expression2 <- expression2[, colnames(expression2) %in% rownames(pheno2)]
expression2 <- as.matrix(expression2)
```

## 3. GSE192444

```{r}
gse3 <- 'GSE192444'
eSet3 <- getGEO('GSE192444')
eSet3 <- eSet3$GSE192444_series_matrix.txt.gz
exp3 <- exprs(eSet3)
phe3 <- pData(eSet3)
feature3 <- fData(eSet3)
```

### Cleaning feature data

```{r}
f3 <- feature3 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f3$'Gene Symbol' <- sapply(strsplit(f3$'Gene Symbol', ' /// '), function(x) x[1])
dim(f3)
```

### Substituting gene probe ID with gene symbol

```{r}
gene_symbols3 <- f3[, c(1, 15)]
expr3 <- as.data.frame(exp3)
expr3[, 'ID'] <- rownames(expr3)
expr3 <- merge(x = expr3, y = gene_symbols3, by.x = 'ID')
dim(expr3)
```

### Adding index

```{r}
rownames(phe3) <- paste(rownames(phe3), '3', sep='_')
colnames(expr3) <- paste(colnames(expr3), '3', sep='_')
dim(phe3)
```

### Removing duplicated gene symbols - keeping the mean gene expression

```{r}
expre3 <- limma::avereps(expr3[, -c(1, ncol(expr3))], ID = expr3$`Gene Symbol`)
expression3 <- as.data.frame(expre3)
dim(expression3)
```

### Filtering of outcome groups

```{r}
phe3$Outcome <- ifelse(str_detect(phe3$`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1`, 'ABMR'), 'ABMR',
                       ifelse(str_detect(phe3$`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1`, 'TCMR'), 'TCMR',
                              ifelse(str_detect(phe3$`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1`, 'NR'), 'Stable', 'Other')))
pheno3 <- phe3 %>% filter(Outcome != 'Other')
table(pheno3$Outcome)

expression3 <- as.data.frame(expression3)
expression3 <- expression3[, colnames(expression3) %in% rownames(pheno3)]
expression3 <- as.matrix(expression3)
```

## 4. GSE124203

```{r}
gse4 <- 'GSE124203'
eSet4 <- getGEO('GSE124203')
eSet4 <- eSet4$GSE124203_series_matrix.txt.gz
exp4 <- exprs(eSet4)
phe4 <- pData(eSet4)
feature4 <- fData(eSet4)
```

### Cleaning feature data

```{r}
f4 <- feature4 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f4$'Gene Symbol' <- sapply(strsplit(f4$'Gene Symbol', ' /// '), function(x) x[1])
dim(f4)
```

### Substituting gene probe ID with gene symbol

```{r}
gene_symbols4 <- f4[, c(1, 15)]
expr4 <- as.data.frame(exp4)
expr4[, 'ID'] <- rownames(expr4)
expr4 <- merge(x = expr4, y = gene_symbols4, by.x = 'ID')
dim(expr4)
```

### Adding index

```{r}
rownames(phe4) <- paste(rownames(phe4), '4', sep='_')
colnames(expr4) <- paste(colnames(expr4), '4', sep='_')
dim(phe4)
```

### Removing duplicated gene symbols - keeping the mean gene expression

```{r}
expre4 <- limma::avereps(expr4[, -c(1, ncol(expr4))], ID = expr4$`Gene Symbol`)
expression4 <- as.data.frame(expre4)
dim(expression4)
```

### Filtering of outcome groups

```{r}
phe4$Outcome <- ifelse(str_detect(phe4$`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1`, 'ABMR'), 'ABMR',
                       ifelse(str_detect(phe4$`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1`, 'TCMR'), 'TCMR',
                              ifelse(str_detect(phe4$`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1`, 'NR'), 'Stable', 'Other')))
pheno4 <- phe4 %>% filter(Outcome != 'Other')
table(pheno4$Outcome)

expression4 <- as.data.frame(expression4)
expression4 <- expression4[, colnames(expression4) %in% rownames(pheno4)]
expression4 <- as.matrix(expression4)
```

## 5. GSE98320

```{r}
gse5 <- 'GSE98320'
eSet5 <- getGEO('GSE98320')
eSet5 <- eSet5$GSE98320_series_matrix.txt.gz
exp5 <- exprs(eSet5)
phe5 <- pData(eSet5)
feature5 <- fData(eSet5)
```

### Cleaning feature data

```{r}
f5 <- feature5 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f5$'Gene Symbol' <- sapply(strsplit(f5$'Gene Symbol', ' /// '), function(x) x[1])
dim(f5)
```

### Substituting gene probe ID with gene symbol

```{r}
gene_symbols5 <- f5[, c(1, 17)]
expr5 <- as.data.frame(exp5)
expr5[, 'ID'] <- rownames(expr5)
expr5 <- merge(x = expr5, y = gene_symbols5, by.x = 'ID')
dim(expr5)
```

### Adding index

```{r}
rownames(phe5) <- paste(rownames(phe5), '5', sep='_')
colnames(expr5) <- paste(colnames(expr5), '5', sep='_')
dim(phe5)
```

### Removing duplicated gene symbols - keeping the mean gene expression

```{r}
expre5 <- limma::avereps(expr5[,-c(1, ncol(expr5))], ID = expr5$`Gene Symbol`)
expression5 <- as.data.frame(expre5)
dim(expression5)
```

### Filtering of outcome groups

```{r}
pheno5 <- phe5 %>% 
  filter(`mmdx:ch1` != '-', `mmdx:ch1` != 'Mixed')%>% 
  mutate(Outcome = case_when(
  `mmdx:ch1` == 'ABMR' ~ 'ABMR',
  `mmdx:ch1` == 'TCMR' ~ 'TCMR',
  `mmdx:ch1` == 'NR' ~ 'Stable',
  TRUE ~ NA_character_
))
table(pheno5$Outcome)

expression5 <- as.data.frame(expression5)
expression5 <- expression5[, colnames(expression5) %in% rownames(pheno5)]
expression5 <- as.matrix(expression5)
```

# Evaluation of datasets, prediction with models

## Selecting the top 75 percentile gene expressions

```{r}
set.seed(3888)
top_75_percentile <- function(eMat) {
  eMat <- as.data.frame(eMat)
  threshold <- quantile(rowSums(eMat), 0.25)
  filtered_eMat <- eMat[rowSums(eMat) > threshold,]
  return (filtered_eMat)
}

expression1 <- top_75_percentile(expression1)
expression2 <- top_75_percentile(expression2)
expression3 <- top_75_percentile(expression3)
expression4 <- top_75_percentile(expression4)
expression5 <- top_75_percentile(expression5)

pheno1 <- pheno1[colnames(expression1),]
pheno2 <- pheno2[colnames(expression2),]
pheno3 <- pheno3[colnames(expression3),]
pheno4 <- pheno4[colnames(expression4),]
pheno5 <- pheno5[colnames(expression5),]
```

## Combining the different datasets

```{r,message=F}
common_genes <- Reduce(intersect, list(rownames(expression1), rownames(expression2), 
                                       rownames(expression3), rownames(expression4), 
                                       rownames(expression5)))

expression1_common <- expression1[common_genes,]
expression2_common <- expression2[common_genes,]
expression3_common <- expression3[common_genes,]
expression4_common <- expression4[common_genes,]
expression5_common <- expression5[common_genes,]

combined_expression <- cbind(expression1_common, expression2_common, 
                             expression3_common, expression4_common, 
                             expression5_common)
dim(combined_expression)
#boxplot(combined_expression)#need normalization and standardization
```

## Adding batch labels

```{r}
# Create a list where each element is a vector of column names for each dataset
batch_labels <- list(
  expression1 = colnames(expression1_common),
  expression2 = colnames(expression2_common),
  expression3 = colnames(expression3_common),
  expression4 = colnames(expression4_common),
  expression5 = colnames(expression5_common)
)
names(batch_labels) <- c("1", "2", "3", "4", "5")
```

## Combined outcomes

```{r}
pheno1 <- pheno1[colnames(expression1_common),]
pheno2 <- pheno2[colnames(expression2_common),]
pheno3 <- pheno3[colnames(expression3_common),]
pheno4 <- pheno4[colnames(expression4_common),]
pheno5 <- pheno5[colnames(expression5_common),]

group1 <- data.frame(Outcome = pheno1$Outcome, row.names = rownames(pheno1))
group2 <- data.frame(Outcome = pheno2$Outcome, row.names = rownames(pheno2))
group3 <- data.frame(Outcome = pheno3$Outcome, row.names = rownames(pheno3))
group4 <- data.frame(Outcome = pheno4$Outcome, row.names = rownames(pheno4))
group5 <- data.frame(Outcome = pheno5$Outcome, row.names = rownames(pheno5))

group <- dplyr::bind_rows(group1, group2, group3, group4, group5)
table(group)
```
### plot stuck plot
```{r}
stuck <- data.frame(
  Dataset = rep(c('GSE36059', 'GSE129166', 'GSE192444', 'GSE124203', 'GSE98320'), each=3),
  Outcome = rep(c('ABMR', 'TCMR', 'Stable'), times=5),
  Count = c(table(pheno1$Outcome), table(pheno2$Outcome), 
            table(pheno3$Outcome), table(pheno4$Outcome), table(pheno5$Outcome))
)
stuck <- stuck %>%
  group_by(Dataset) %>%
  mutate(Proportion = Count / sum(Count)) %>%
  ungroup()
ggplot(stuck, aes(x = Dataset, y = Count, fill = Outcome)) +
  geom_bar(stat = 'identity', position = 'stack') +
  labs(title = "Proportion of Outcome Types by Dataset",
       x = "Dataset",
       y = "Proportion of Samples",
       fill = "Outcome Type") +
  theme_minimal()+
  scale_fill_manual(values = c('ABMR' = 'navyblue', 'TCMR' = 'lightblue', "Stable" = 'grey'))

```
## Balancing of outcome groups and normalising expression intensities

```{r}
set.seed(3888)

# Aim: ABMR = TCMR = Stable = 241
# Picking random samples
ABMR_samples <- sample(colnames(combined_expression)[group == 'ABMR'], 241)
set.seed(3888)
TCMR_samples <- sample(colnames(combined_expression)[group == 'TCMR'], 241)
set.seed(3888)
Stable_samples <- sample(colnames(combined_expression)[group == 'Stable'], 241)
selected_samples <- c(ABMR_samples, TCMR_samples, Stable_samples)

Combined_Expression<- log2(combined_expression[, selected_samples]+1)
#boxplot(cbind_expression)
#normalization + standardization
#Combined_Expression <- normalizeBetweenArrays(cbind_expression)
#boxplot(Combined_Expression)

filter_batches <- function(batches, samples) {
  lapply(batches, function(batch) batch[batch %in% samples])
}

batch_labels <- filter_batches(batch_labels, selected_samples)

group <- group[selected_samples, , drop = FALSE]
batch_info <- apply(sapply(batch_labels, FUN = `%in%`, x = rownames(group)), MARGIN = 1, FUN = which)

table(group$Outcome)
dim(Combined_Expression) # n = 241 * 3 = 723
# print(lapply(batch_labels, length))
```
### plot stack plot
```{r}
group_balance <- group %>% filter(rownames(group) %in% colnames(Combined_Expression))
Sample <- rownames(group_balance)
group_balance <- group_balance %>%
  mutate(Dataset = case_when(
    grepl("_1", Sample) ~ "GSE36059",
    grepl("_2", Sample) ~ "GSE129166",
    grepl("_3", Sample) ~ "GSE192444",
    grepl("_4", Sample) ~ "GSE124203",
    grepl("_5", Sample) ~ "GSE98320"
  ))
group_balance <- group_balance%>% 
  group_by(Dataset)%>% 
  mutate(Count = length(Outcome))%>% 
  mutate(Proportion = Count / sum(Count))%>%
  ungroup()
ggplot(group_balance, aes(x = Dataset, y = Count, fill = Outcome)) +
  geom_bar(stat = 'identity', position = 'stack') +
  labs(title = "Proportion of Outcome Types by Dataset after Balancing",
       x = "Dataset",
       y = "Proportion of Samples",
       fill = "Outcome Type") +
  theme_minimal()+
  scale_fill_manual(values = c('ABMR' = 'navyblue', 'TCMR' = 'lightblue', "Stable" = 'grey'))
table(group %>% filter(rownames(group) %in% colnames(Combined_Expression)))
```

## taking 70% of the original data to be the training, 30% to tbe the final testing
```{r}
set.seed(3888)
n<- ncol(Combined_Expression)
idx <- sample(n)
set.seed(3888)
split_idx<- floor(0.3*n)

eMat_30 <-Combined_Expression[, idx[1:split_idx]]
eMat_70 <- Combined_Expression[, idx[(split_idx + 1):n]]
dim(eMat_30)
dim(eMat_70)
group_70 <- group %>% filter(rownames(group) %in% colnames(eMat_70))
table(group %>% filter(rownames(group) %in% colnames(eMat_70)))
```
### plot stack bar plot
```{r}
Sample_split <- rownames(group_70)
group_70 <- group_70 %>%
  mutate(Dataset = case_when(
    grepl("_1", Sample_split) ~ "GSE36059",
    grepl("_2", Sample_split) ~ "GSE129166",
    grepl("_3", Sample_split) ~ "GSE192444",
    grepl("_4", Sample_split) ~ "GSE124203",
    grepl("_5", Sample_split) ~ "GSE98320"
  ))
group_70 <- group_70%>% 
  group_by(Dataset)%>% 
  mutate(Count = length(Outcome))%>% 
  mutate(Proportion = Count / sum(Count))%>%
  ungroup()
ggplot(group_70, aes(x = Dataset, y = Count, fill = Outcome)) +
  geom_bar(stat = 'identity', position = 'stack') +
  labs(title = "Proportion of Outcome Types by Dataset after Balancing",
       x = "Dataset",
       y = "Proportion of Samples",
       fill = "Outcome Type") +
  theme_minimal()+
  scale_fill_manual(values = c('ABMR' = 'navyblue', 'TCMR' = 'lightblue', "Stable" = 'grey'))
table(group %>% filter(rownames(group) %in% colnames(eMat_70)))
```

### adjusting batch labels
```{r}
batch_info_70 <- apply(sapply(batch_labels, FUN = `%in%`, x = colnames(eMat_70)), MARGIN = 1, FUN = which)
```

## Exploring data with PCA - before batch correction

```{r}
set.seed(3888)

gse_pca <- prcomp(t(eMat_70))
df_toplot <- data.frame(Outcome = group[colnames(eMat_70),],
                        pc1 = gse_pca$x[,1], pc2 = gse_pca$x[,2],
                        batch = batch_info_70)

ggplot(df_toplot, aes(x = pc1, y = pc2, color = Outcome)) + 
  geom_point(size = 1.5) + labs(colour = "Outcome") +
  theme_minimal()

ggplot(df_toplot, aes(x = pc1, y = pc2, color = factor(batch))) + 
  geom_point(size = 1.5) + labs(colour = "Batch") +
  theme_minimal()
```

## ComBat batch correction

```{r, echo=FALSE,message=FALSE}
set.seed(3888)
MOD <- model.matrix(~factor(group[colnames(eMat_70),]))
combat_eMat_70 <- sva::ComBat(dat = eMat_70, batch = batch_info_70, mod = MOD, par.prior = TRUE, prior.plots = FALSE) 
dim(combat_eMat_70)
boxplot(combat_eMat_70)
```

## Exploring data with PCA - after batch correction

```{r}
set.seed(3888)

gse_pca <- prcomp(t(combat_eMat_70))
df_toplot <- data.frame(Outcome = group[colnames(combat_eMat_70),],
                        pc1 = gse_pca$x[,1], pc2 = gse_pca$x[,2],
                        batch = as.factor(batch_info_70))

ggplot(df_toplot, aes(x = pc1, y = pc2, color = Outcome)) + 
  geom_point(size = 1.5) + labs(colour = "Outcome") +
  theme_minimal()

ggplot(df_toplot, aes(x = pc1, y = pc2, color = batch)) + 
  geom_point(size = 1.5) + labs(colour = "Batch") +
  theme_minimal()
```
#### Export my data set
```{r}
#write.csv(eMat_70,file="/Users/lululiya/Desktop/code/eMat_70.csv",quote=F,row.names = T)
#write.csv(eMat_30,file="/Users/lululiya/Desktop/code/eMat_30.csv",quote=F,row.names = T)
#write.csv(combat_eMat_70,file="/Users/lululiya/Desktop/code/combat_eMat_70.csv",quote=F,row.names = T)
#write.csv(group,file="/Users/lululiya/Desktop/code/Outcome.csv",quote=F,row.names = T)
```


## Model evaluation with cross-validation
### Limma top genes selectoin function
```{r}
set.seed(3888)
limma_top_genes <- function(eMat, outcome, n) {
  group_factor = factor(outcome)
  design_matrix <- model.matrix(~ 0 + group_factor)
  colnames(design_matrix) <- levels(group_factor)
      
  contrast_matrix <- makeContrasts(
    ABMR_vs_Stable_and_TCMR = ABMR - (1 / 2 * (Stable + TCMR)),
    TCMR_vs_Stable_and_ABMR = TCMR - (1 / 2 * (Stable + ABMR)),
    Stable_vs_Rejection = Stable - (1 / 2 * (ABMR + TCMR)),
    levels = colnames(design_matrix)
  )
      
  fit <- lmFit(eMat, design = design_matrix)
  fit <- contrasts.fit(fit, contrasts = contrast_matrix)
  fit <- eBayes(fit)
  
  tT_ABMR <- topTable(fit, n = n, coef = 1)
  tT_TCMR <- topTable(fit, n = n, coef = 2)
  tT_Stable <- topTable(fit, n = n, coef = 3)

  top_genes <- unique(c(rownames(tT_ABMR), rownames(tT_TCMR), rownames(tT_Stable)))
  
  return (top_genes)
}
```
### cross-validaton
```{r}
set.seed(3888)

X = t(eMat_70)
y = factor(group[colnames(eMat_70),])

cvK = 5   
n_sim = 20

cv_accuracy30_rf = cv_accuracy50_rf = cv_accuracy110_rf = cv_accuracy80_rf=numeric(n_sim)
cv_accuracy30_knn2 = cv_accuracy50_knn2 = cv_accuracy110_knn2 = cv_accuracy80_knn2 =numeric(n_sim)
cv_accuracy30_knn5 = cv_accuracy50_knn5 = cv_accuracy110_knn5 = cv_accuracy80_knn5=numeric(n_sim)
cv_accuracy30_svm = cv_accuracy50_svm = cv_accuracy110_svm = cv_accuracy80_svm =numeric(n_sim)

for (i in 1:n_sim) {
  cvSets = cvFolds(nrow(X), cvK)

  cv_accuracy_folds30_rf = cv_accuracy_folds50_rf = cv_accuracy_folds110_rf = cv_accuracy_folds80_rf=numeric(cvK)
  cv_accuracy_folds30_knn2 = cv_accuracy_folds50_knn2 = cv_accuracy_folds110_knn2 = cv_accuracy_folds80_knn2=numeric(cvK)
  cv_accuracy_folds30_knn5 = cv_accuracy_folds50_knn5 = cv_accuracy_folds110_knn5 = cv_accuracy_folds80_knn5=numeric(cvK)
  cv_accuracy_folds30_svm = cv_accuracy_folds50_svm = cv_accuracy_folds110_svm = cv_accuracy_folds80_svm=numeric(cvK)
  
  
  for (j in 1:cvK) {
    test_id = cvSets$subsets[cvSets$which == j]
    X_train = X[-test_id,]
    X_test = X[test_id,]
    y_train = y[-test_id]
    y_test = y[test_id]
    set.seed(3888)
    # Applying batch correction to the training data
    batch_train <- apply(sapply(batch_labels, FUN = `%in%`, x = rownames(X_train)), MARGIN = 1, FUN = which)
    mod <- model.matrix(~factor(group[rownames(X_train),]))
    X_train <- t(sva::ComBat(dat = t(X_train), batch = batch_train, mod = mod, par.prior = TRUE, prior.plots = FALSE))
    
    # Feature selection with Limma
    DE_genes30 <- limma_top_genes(t(X_train), y_train, 10)
    DE_genes50 <- limma_top_genes(t(X_train), y_train, 17)
    DE_genes80 <- limma_top_genes(t(X_train), y_train, 27)
    DE_genes110 <- limma_top_genes(t(X_train), y_train, 37)
    
    X_train30 = X_train[,DE_genes30]
    X_test30 <- X_test[,DE_genes30]
    
    X_train50 = X_train[,DE_genes50]
    X_test50 = X_test[,DE_genes50]
    
    X_train80 = X_train[,DE_genes80]
    X_test80 = X_test[,DE_genes80]
    
    X_train110 = X_train[,DE_genes110]
    X_test110 = X_test[,DE_genes110]
    
    
    # Random Forest
    rf_fit30 = randomForest(x = X_train30, y = y_train)
    predictions_rf30 = predict(rf_fit30, X_test30)
    cv_accuracy_folds30_rf[j] = mean(y_test == predictions_rf30)
    
    rf_fit50 = randomForest(x = X_train50, y = y_train)
    predictions_rf50 = predict(rf_fit50, X_test50)
    cv_accuracy_folds50_rf[j] = mean(y_test == predictions_rf50)
    
    rf_fit80 = randomForest(x = X_train80, y = y_train)
    predictions_rf80 = predict(rf_fit80, X_test80)
    cv_accuracy_folds80_rf[j] = mean(y_test == predictions_rf80)
    
    rf_fit110 = randomForest(x = X_train110, y = y_train)
    predictions_rf110 = predict(rf_fit110, X_test110)
    cv_accuracy_folds110_rf[j] = mean(y_test == predictions_rf110)
    
    
    # SVM
    svm_fit30 = svm(X_train30, y_train)
    predictions_svm30 = predict(svm_fit30, X_test30)
    cv_accuracy_folds30_svm[j] = mean(y_test == predictions_svm30)
    
    svm_fit50 = svm(X_train50, y_train)
    predictions_svm50 = predict(svm_fit50, X_test50)
    cv_accuracy_folds50_svm[j] = mean(y_test == predictions_svm50)
    
    svm_fit80 = svm(X_train80, y_train)
    predictions_svm80 = predict(svm_fit80, X_test80)
    cv_accuracy_folds80_svm[j] = mean(y_test == predictions_svm80)
    
    svm_fit110 = svm(X_train110, y_train)
    predictions_svm110 = predict(svm_fit110, X_test110)
    cv_accuracy_folds110_svm[j] = mean(y_test == predictions_svm110)
    
    
    # KNN (k = 2)
     knn2_fit30 = class::knn(train = X_train30, test = X_test30, cl = y_train, k = 2)
     cv_accuracy_folds30_knn2[j] = mean(y_test == knn2_fit30)
     
     knn2_fit50 = class::knn(train = X_train50, test = X_test50, cl = y_train, k = 2)
     cv_accuracy_folds50_knn2[j] = mean(y_test == knn2_fit50)
     
     knn2_fit80 = class::knn(train = X_train80, test = X_test80, cl = y_train, k = 2)
     cv_accuracy_folds80_knn2[j] = mean(y_test == knn2_fit80)
     
     knn2_fit110 = class::knn(train = X_train110, test = X_test110, cl = y_train, k = 2)
     cv_accuracy_folds110_knn2[j] = mean(y_test == knn2_fit110)
    
     
    # KNN (k = 5)
     knn5_fit30 = class::knn(train = X_train30, test = X_test30, cl = y_train, k = 5)
     cv_accuracy_folds30_knn5[j] = mean(y_test == knn5_fit30)
     
     knn5_fit50 = class::knn(train = X_train50, test = X_test50, cl = y_train, k = 5)
     cv_accuracy_folds50_knn5[j] = mean(y_test == knn5_fit50)
     
     knn5_fit80 = class::knn(train = X_train80, test = X_test80, cl = y_train, k = 5)
     cv_accuracy_folds80_knn5[j] = mean(y_test == knn5_fit80)
     
     knn5_fit110 = class::knn(train = X_train110, test = X_test110, cl = y_train, k = 5)
     cv_accuracy_folds110_knn5[j] = mean(y_test == knn5_fit110)
  }
  
  cv_accuracy30_rf[i] = mean(cv_accuracy_folds30_rf)
  cv_accuracy50_rf[i] = mean(cv_accuracy_folds50_rf)
  cv_accuracy80_rf[i] = mean(cv_accuracy_folds80_rf)
  cv_accuracy110_rf[i] = mean(cv_accuracy_folds110_rf)
  
  cv_accuracy30_svm[i] = mean(cv_accuracy_folds30_svm)
  cv_accuracy50_svm[i] = mean(cv_accuracy_folds50_svm)
  cv_accuracy80_svm[i] = mean(cv_accuracy_folds80_svm)
  cv_accuracy110_svm[i] = mean(cv_accuracy_folds110_svm)
  
  cv_accuracy30_knn2[i] = mean(cv_accuracy_folds30_knn2)
  cv_accuracy50_knn2[i] = mean(cv_accuracy_folds50_knn2)
  cv_accuracy80_knn2[i] = mean(cv_accuracy_folds80_knn2)
  cv_accuracy110_knn2[i] = mean(cv_accuracy_folds110_knn2)
 
  cv_accuracy30_knn5[i] = mean(cv_accuracy_folds30_knn5)
  cv_accuracy50_knn5[i] = mean(cv_accuracy_folds50_knn5)
  cv_accuracy80_knn5[i] = mean(cv_accuracy_folds80_knn5)
  cv_accuracy110_knn5[i] = mean(cv_accuracy_folds110_knn5)
}

 boxplot(cv_accuracy30_rf, cv_accuracy50_rf, cv_accuracy80_rf,cv_accuracy110_rf,
         cv_accuracy30_svm, cv_accuracy50_svm, cv_accuracy80_svm,cv_accuracy110_svm,
         cv_accuracy30_knn2, cv_accuracy50_knn2, cv_accuracy80_knn2,cv_accuracy110_knn2,
         cv_accuracy30_knn5, cv_accuracy50_knn5,cv_accuracy80_knn5, cv_accuracy110_knn5,
        names = c("RF - 30", "RF - 50",'RF - 80', "RF - 110", 'SVM -30','SVM -50','SVM - 80','SVM -110',"KNN2 - 30", "KNN2 - 50", 'KNN - 80',"KNN2 - 110", "KNN5 - 30", "KNN5 - 50", 'KNN - 80',"KNN5 - 110"),las=2,cex.axis = 0.65)

```

#### Evaluation of cv

```{r}
comparision <- data.frame(Model =c('rf-30','rf-50','rf-80','rf-110','knn5-30','knn5-50','knn5-80','knn5-110'),
                          Mean=c(mean(cv_accuracy30_rf),
                                 mean(cv_accuracy50_rf),
                                 mean(cv_accuracy80_rf),
                                 mean(cv_accuracy110_rf),
                                 mean(cv_accuracy50_knn5),
                                 mean(cv_accuracy30_knn5),
                                 mean(cv_accuracy80_knn5),
                                 mean(cv_accuracy110_knn5)),
                          Variance=c(var(cv_accuracy30_rf),
                                     var(cv_accuracy50_rf),
                                 var(cv_accuracy80_rf),
                                 var(cv_accuracy110_rf),
                                 var(cv_accuracy30_knn5),
                                var(cv_accuracy50_knn5),
                                 var(cv_accuracy80_knn5),
                                var(cv_accuracy110_knn5)))

#comparision
#accuracies all similar
comparision%>% arrange(Variance)
#so pick the one thathas smaller variance and smaller number of features
#then rf-50
```


# Final Model
## 50 top gens
### Limma top 50 selection
```{r}
set.seed(3888)
group_factor_final = factor(group[colnames(combat_eMat_70),])
design_matrix_final <- model.matrix(~ 0 + group_factor_final)
colnames(design_matrix_final) <- levels(group_factor_final)
      
contrast_matrix_final <- makeContrasts(
    ABMR_vs_Stable_and_TCMR = ABMR - (1 / 2 * (Stable + TCMR)),
    TCMR_vs_Stable_and_ABMR = TCMR - (1 / 2 * (Stable + ABMR)),
    Stable_vs_Rejection = Stable - (1 / 2 * (ABMR + TCMR)),
    levels = colnames(design_matrix_final)
  )

Fit <- lmFit(combat_eMat_70, design = design_matrix_final)
Fit <- contrasts.fit(Fit, contrasts = contrast_matrix_final)
Fit <- eBayes(Fit)
  
tT_ABMR_final <- topTable(Fit, n = 17, coef = 1)%>% arrange('P.Value')
tT_TCMR_final <- topTable(Fit, n = 17, coef = 2)%>% arrange('P.Value')
tT_Stable_final <- topTable(Fit, n = 17, coef = 3)%>% arrange('P.Value')

top_genes_final <- unique(c(rownames(tT_ABMR_final), rownames(tT_TCMR_final), rownames(tT_Stable_final)))#49
length(top_genes_final)
```

## Final model (knn 50) prepareation
```{r}
set.seed(3888)
#eMat_30 = testing data
#combat_eMat_70 = training data
#top_genes_final = top genes

X_test_final_model <- eMat_30[rownames(eMat_30) %in% top_genes_final, ]
#X_test_final <- t(normalizeBetweenArrays(X_test_final))
X_test_final_model <- t(X_test_final_model)
X_train_final_model <- t(combat_eMat_70[rownames(combat_eMat_70) %in% top_genes_final, ])
y_train_final_model <- factor(group[rownames(X_train_final_model),])
y_test_final_model <-factor(group[rownames(X_test_final_model),])
```
### rf -50
```{r}
set.seed(3888)
rf_fit_final <- randomForest(x = X_train_final_model, y = y_train_final_model)
predicted_final_model <- predict(rf_fit_final, newdata=X_test_final_model)

c <- confusionMatrix(as.factor(predicted_final_model), as.factor(y_test_final_model))
c
```

## Evaluation of the final model
```{r,message=FALSE,warning=FALSE}
response_abmr <- factor(ifelse(y_test_final_model == 'ABMR', 'ABMR', 'Not ABMR'))
response_tcmr <- factor(ifelse(y_test_final_model == 'TCMR', 'TCMR', 'Not TCMR'))
response_stable <- factor(ifelse(y_test_final_model == 'Stable', 'Stable', 'Not Stable'))

roc_abmr <- roc(response = response_abmr, 
                predictor = prob_predictions[, 'ABMR'])
roc_tcmr <- roc(response = response_tcmr, 
                predictor = prob_predictions[, 'TCMR'])
roc_stable <- roc(response = response_stable, 
                  predictor = prob_predictions[, 'Stable'])


paste('abmr:',round(roc_abmr$auc,3))
paste('tcmr:',round(roc_tcmr$auc,3))
paste('stable:',round(roc_stable$auc,3))

plot(roc_abmr, col = "red", main = 'ROC Curves for Each Class')
plot(roc_tcmr, col = "blue", add = TRUE)
plot(roc_stable, col = "green", add = TRUE)
legend('bottomright', legend = c('ABMR vs. Others', 'TCMR vs. Others', 'Stable vs. Others'), 
       col = c('red', 'blue', 'green'), lwd = 2)

legend("topright", inset=c(-0.005, 0), # Adjust the inset to move the legend inside the plot area
       legend = c(paste("AUC ABMR: ", round(roc_abmr$auc,3)),
                  paste("AUC TCMR: ", round(roc_tcmr$auc,3)),
                  paste("AUC Stable: ", round(roc_stable$auc,3))),
       title = "AUC Values", 
       bty = "n", # No box around the legend
       text.col = c("red", "blue", "green"))

```