---
title: "DATA3888 Interdisciplinary Project"
author: 'Kidney20'
output:
  html_document:
    fig_caption: yes
    self_contained: yes
    theme: flatly
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
---

# Loading required packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(caret)
library(cvTools)
library(dplyr)
library(e1071)
library(GEOquery)
library(ggplot2)
library(glmnet)
library(limma)
library(pheatmap)
library(randomForest)
library(stringr)
library(sva)
library(tidyverse)

rm(list = ls())
```


# Loading datasets and cleaning

## 1. GSE36059

```{r}
gse1 <- 'GSE36059'
eSet1 <- getGEO('GSE36059')
eSet1 <- eSet1$GSE36059_series_matrix.txt.gz
exp1 <- exprs(eSet1)
phe1 <- pData(eSet1)
feature1 <- fData(eSet1)
```

### Cleaning feature data

```{r}
f1 <- feature1 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f1$'Gene Symbol' <- sapply(strsplit(f1$'Gene Symbol', ' /// '), function(x) x[1])
```

### Substituting gene probe ID with gene symbol

```{r}
gene_symbols1 <- f1[, c(1, 11)]
expr1 <- as.data.frame(exp1)
expr1[, 'ID'] <- rownames(expr1)
expr1 <- merge(x = expr1, y = gene_symbols1, by.x = 'ID')
```

### Adding index

```{r}
rownames(phe1) <- paste(rownames(phe1), '1', sep='_')
colnames(expr1) <- paste(colnames(expr1), '1', sep='_')
```

### Removing duplicated gene symbols - keeping the mean gene expression

```{r}
expre1 <- limma::avereps(expr1[, -c(1, ncol(expr1))], ID = expr1$`Gene Symbol`)
expre1 <- as.data.frame(expre1)
expression1 <- expre1[, -c(1, ncol(expre1))]
dim(expression1)
```

### Filtering of outcome groups

```{r}
phe1$Outcome <- ifelse(str_detect(phe1$title, 'non-rejecting'), 'Stable',
                       ifelse(str_detect(phe1$title, 'ABMR'), 'ABMR',
                              ifelse(str_detect(phe1$title, 'TCMR'), 'TCMR', 'Other')))
pheno1 <- phe1 %>% filter(Outcome != 'Other')
table(pheno1$Outcome)

expression1 <- as.data.frame(expression1)
expression1 <- expression1[, colnames(expression1) %in% rownames(pheno1)]
expression1 <- as.matrix(expression1)
```


## 2. GSE129166

```{r}
gse2 <- 'GSE129166'
eSet2 <- getGEO('GSE129166')
eSet2 <- eSet2$GSE129166_series_matrix.txt.gz
exp2 <- exprs(eSet2)
phe2 <- pData(eSet2)
feature2 <- fData(eSet2)
```

### Filtering to biopsy data only

```{r}
phe2 <- phe2 %>% filter(str_detect(title, 'BIOPSY'))
exp2 <- exp2[, rownames(phe2)]
```

### Cleaning feature data

```{r}
f2 <- feature2 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f2$'Gene Symbol' <- sapply(strsplit(f2$'Gene Symbol', ' /// '), function(x) x[1])
```

### Substituting gene probe ID with gene symbol

```{r}
gene_symbols2 <- f2[, c(1, 11)]
expr2 <- as.data.frame(exp2)
expr2[, 'ID'] <- rownames(expr2)
expr2 <- merge(x = expr2, y = gene_symbols2, by.x = 'ID')
```

### Adding index

```{r}
rownames(phe2) <- paste(rownames(phe2), '2', sep='_')
colnames(expr2) <- paste(colnames(expr2), '2', sep='_')
```

### Removing duplicated gene symbols - keeping the mean gene expression

```{r}
expre2 <- limma::avereps(expr2[, -c(1, ncol(expr2))], ID = expr2$`Gene Symbol`)
expre2 <- as.data.frame(expre2)
expression2 <- expre2[, -c(1, ncol(expre2))]
dim(expression2)
```

### Filtering of outcome groups

```{r}
pheno2 <- phe2 %>% mutate(
  ABMR_Status = ifelse(`abmr (no:ch1` == '0_Yes:1): 1', 'ABMR', 'ABMR Stable'),
  TCMR_Status = ifelse(`tcmr (no:ch1` == '0_borderline:1_TCMR:2): 1', 'TCMR', 'TCMR Stable')
  ) %>%
    filter(!(ABMR_Status == 'ABMR' & TCMR_Status == 'TCMR')) %>%
  mutate(
    Outcome = case_when(
      ABMR_Status == 'ABMR Stable' & TCMR_Status == 'TCMR Stable' ~ 'Stable',
      ABMR_Status == 'ABMR Stable' ~ 'TCMR',
      TCMR_Status == 'TCMR Stable' ~ 'ABMR',
      TRUE ~ NA_character_
    )
  )
table(pheno2$Outcome)

expression2 <- as.data.frame(expression2)
expression2 <- expression2[, colnames(expression2) %in% rownames(pheno2)]
expression2 <- as.matrix(expression2)
```


## 3. GSE192444

```{r}
gse3 <- 'GSE192444'
eSet3 <- getGEO('GSE192444')
eSet3 <- eSet3$GSE192444_series_matrix.txt.gz
exp3 <- exprs(eSet3)
phe3 <- pData(eSet3)
feature3 <- fData(eSet3)
```

### Cleaning feature data

```{r}
f3 <- feature3 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f3$'Gene Symbol' <- sapply(strsplit(f3$'Gene Symbol', ' /// '), function(x) x[1])
```

### Substituting gene probe ID with gene symbol

```{r}
gene_symbols3 <- f3[, c(1, 15)]
expr3 <- as.data.frame(exp3)
expr3[, 'ID'] <- rownames(expr3)
expr3 <- merge(x = expr3, y = gene_symbols3, by.x = 'ID')
```

### Adding index

```{r}
rownames(phe3) <- paste(rownames(phe3), '3', sep='_')
colnames(expr3) <- paste(colnames(expr3), '3', sep='_')
```

### Removing duplicated gene symbols - keeping the mean gene expression

```{r}
expre3 <- limma::avereps(expr3[, -c(1, ncol(expr3))], ID = expr3$`Gene Symbol`)
expre3 <- as.data.frame(expre3)
expression3 <- expre3[, -c(1,ncol(expre3))]
dim(expression3)
```

### Filtering of outcome groups

```{r}
phe3$Outcome <- ifelse(str_detect(phe3$`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1`, 'ABMR'), 'ABMR',
                       ifelse(str_detect(phe3$`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1`, 'TCMR'), 'TCMR',
                              ifelse(str_detect(phe3$`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1`, 'NR'), 'Stable', 'Other')))
pheno3 <- phe3 %>% filter(Outcome != 'Other')
table(pheno3$Outcome)

expression3 <- as.data.frame(expression3)
expression3 <- expression3[, colnames(expression3) %in% rownames(pheno3)]
expression3 <- as.matrix(expression3)
```


## 4. GSE124203

```{r}
gse4 <- 'GSE124203'
eSet4 <- getGEO('GSE124203')
eSet4 <- eSet4$GSE124203_series_matrix.txt.gz
exp4 <- exprs(eSet4)
phe4 <- pData(eSet4)
feature4 <- fData(eSet4)
```

### Cleaning feature data

```{r}
f4 <- feature4 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f4$'Gene Symbol' <- sapply(strsplit(f4$'Gene Symbol', ' /// '), function(x) x[1])
```

### Substituting gene probe ID with gene symbol

```{r}
gene_symbols4 <- f4[, c(1, 15)]
expr4 <- as.data.frame(exp4)
expr4[, 'ID'] <- rownames(expr4)
expr4 <- merge(x = expr4, y = gene_symbols4, by.x = 'ID')
```

### Adding index

```{r}
rownames(phe4) <- paste(rownames(phe4), '4', sep='_')
colnames(expr4) <- paste(colnames(expr4), '4', sep='_')
```

### Removing duplicated gene symbols - keeping the mean gene expression

```{r}
expre4 <- limma::avereps(expr4[, -c(1, ncol(expr4))], ID = expr4$`Gene Symbol`)
expre4 <- as.data.frame(expre4)
expression4 <- expre4[, -c(1, ncol(expre4))]
dim(expression4)
```

### Filtering of outcome groups

```{r}
phe4$Outcome <- ifelse(str_detect(phe4$`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1`, 'ABMR'), 'ABMR',
                       ifelse(str_detect(phe4$`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1`, 'TCMR'), 'TCMR',
                              ifelse(str_detect(phe4$`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1`, 'NR'), 'Stable', 'Other')))
pheno4 <- phe4 %>% filter(Outcome != 'Other')
table(pheno4$Outcome)

expression4 <- as.data.frame(expression4)
expression4 <- expression4[, colnames(expression4) %in% rownames(pheno4)]
expression4 <- as.matrix(expression4)
```


## 5. GSE98320

```{r}
gse5 <- 'GSE98320'
eSet5 <- getGEO('GSE98320')
eSet5 <- eSet5$GSE98320_series_matrix.txt.gz
exp5 <- exprs(eSet5)
phe5 <- pData(eSet5)
feature5 <- fData(eSet5)
```

### Cleaning feature data

```{r}
f5 <- feature5 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f5$'Gene Symbol' <- sapply(strsplit(f5$'Gene Symbol', ' /// '), function(x) x[1])
```

### Substituting gene probe ID with gene symbol

```{r}
gene_symbols5 <- f5[, c(1, 17)]
expr5 <- as.data.frame(exp5)
expr5[, 'ID'] <- rownames(expr5)
expr5 <- merge(x = expr5, y = gene_symbols5, by.x = 'ID')
```

### Adding index

```{r}
rownames(phe5) <- paste(rownames(phe5), '5', sep='_')
colnames(expr5) <- paste(colnames(expr5), '5', sep='_')
```

### Removing duplicated gene symbols - keeping the mean gene expression

```{r}
expre5 <- limma::avereps(expr5[,-c(1, ncol(expr5))], ID = expr5$`Gene Symbol`)
expre5 <- as.data.frame(expre5)
expression5 <- expre5[, -c(1, ncol(expre5))]
dim(expression5)
```

### Filtering of outcome groups

```{r}
pheno5 <- phe5 %>%
  mutate(
    Outcome = ifelse(`d96:ch1` == 'ABMR','ABMR',
                     ifelse(`d96:ch1`=='TCMR', 'TCMR', NA))) %>%
  filter(!is.na(Outcome))
table(pheno5$Outcome)

expression5 <- as.data.frame(expression5)
expression5 <- expression5[, colnames(expression5) %in% rownames(pheno5)]
expression5 <- as.matrix(expression5)
```


# Evaluation of datasets, prediction with models

## Selecting the top 75 percentile gene expressions

```{r}
top_75_percentile <- function(eMat) {
  eMat <- as.data.frame(eMat)
  threshold <- quantile(rowSums(eMat), 0.25)
  filtered_eMat <- eMat[rowSums(eMat) > threshold,]
  return (filtered_eMat)
}

expression1 <- top_75_percentile(expression1)
expression2 <- top_75_percentile(expression2)
expression3 <- top_75_percentile(expression3)
expression4 <- top_75_percentile(expression4)
expression5 <- top_75_percentile(expression5)

pheno1 <- pheno1[colnames(expression1),]
pheno2 <- pheno2[colnames(expression2),]
pheno3 <- pheno3[colnames(expression3),]
pheno4 <- pheno4[colnames(expression4),]
pheno5 <- pheno5[colnames(expression5),]
```


## Combining the different datasets

```{r,message=F}
common_genes <- Reduce(intersect, list(rownames(expression1), rownames(expression2), 
                                       rownames(expression3), rownames(expression4), 
                                       rownames(expression5)))

expression1_common <- expression1[common_genes,]
expression2_common <- expression2[common_genes,]
expression3_common <- expression3[common_genes,]
expression4_common <- expression4[common_genes,]
expression5_common <- expression5[common_genes,]

combined_expression <- cbind(expression1_common, expression2_common, 
                             expression3_common, expression4_common, 
                             expression5_common)
dim(combined_expression)
```


## Adding batch labels

```{r}
# Create a list where each element is a vector of column names for each dataset
batch_labels <- list(
  expression1 = colnames(expression1_common),
  expression2 = colnames(expression2_common),
  expression3 = colnames(expression3_common),
  expression4 = colnames(expression4_common),
  expression5 = colnames(expression5_common)
)
names(batch_labels) <- c("1", "2", "3", "4", "5")
```


## Combined outcomes

```{r}
pheno1 <- pheno1[colnames(expression1_common),]
pheno2 <- pheno2[colnames(expression2_common),]
pheno3 <- pheno3[colnames(expression3_common),]
pheno4 <- pheno4[colnames(expression4_common),]
pheno5 <- pheno5[colnames(expression5_common),]

group1 <- data.frame(Outcome = pheno1$Outcome, row.names = rownames(pheno1))
group2 <- data.frame(Outcome = pheno2$Outcome, row.names = rownames(pheno2))
group3 <- data.frame(Outcome = pheno3$Outcome, row.names = rownames(pheno3))
group4 <- data.frame(Outcome = pheno4$Outcome, row.names = rownames(pheno4))
group5 <- data.frame(Outcome = pheno5$Outcome, row.names = rownames(pheno5))

group <- dplyr::bind_rows(group1, group2, group3, group4, group5)
table(group)
```


## Balancing of outcome groups and normalising expression intensities

```{r}
set.seed(3888)

# Aim: ABMR = TCMR = Stable = 302
# Picking random samples
ABMR_samples <- sample(colnames(combined_expression)[group == 'ABMR'], 302)
TCMR_samples <- sample(colnames(combined_expression)[group == 'TCMR'], 302)
Stable_samples <- sample(colnames(combined_expression)[group == 'Stable'], 302)
selected_samples <- c(ABMR_samples, TCMR_samples, Stable_samples)

combined_expression <- combined_expression[, selected_samples] 
combined_expression <- normalizeBetweenArrays(combined_expression)

filter_batches <- function(batches, samples) {
  lapply(batches, function(batch) batch[batch %in% samples])
}

batch_labels <- filter_batches(batch_labels, selected_samples)

group <- group[selected_samples, , drop = FALSE]
batch_info <- apply(sapply(batch_labels, FUN = `%in%`, x = rownames(group)), MARGIN = 1, FUN = which)

table(group)
dim(combined_expression) # n = 302 * 3 = 906
# print(lapply(batch_labels, length))
```


## Exploring data with PCA - before batch correction

```{r}
set.seed(3888)

gse_pca <- prcomp(t(combined_expression))
df_toplot <- data.frame(Outcome = group$Outcome,
                        pc1 = gse_pca$x[,1], pc2 = gse_pca$x[,2],
                        batch = batch_info)

ggplot(df_toplot, aes(x = pc1, y = pc2, color = Outcome)) + 
  geom_point(size = 1.5) + labs(colour = "Outcome") +
  theme_minimal()

ggplot(df_toplot, aes(x = pc1, y = pc2, color = batch)) + 
  geom_point(size = 1.5) + labs(colour = "Batch") +
  theme_minimal()
```


## ComBat-seq batch correction

```{r, echo=FALSE}
mod <- model.matrix(~1, data = data.frame(ones = rep(1, ncol(combined_expression))))
combat_eMat <- sva::ComBat(dat = combined_expression, batch = batch_info, mod = mod, par.prior = TRUE, prior.plots = FALSE)
dim(combat_eMat)
```


## Exploring data with PCA - after batch correction

```{r}
set.seed(3888)

gse_pca <- prcomp(t(combat_eMat))
df_toplot <- data.frame(Outcome = group$Outcome,
                        pc1 = gse_pca$x[,1], pc2 = gse_pca$x[,2],
                        batch = as.factor(batch_info))

ggplot(df_toplot, aes(x = pc1, y = pc2, color = Outcome)) + 
  geom_point(size = 1.5) + labs(colour = "Outcome") +
  theme_minimal()

ggplot(df_toplot, aes(x = pc1, y = pc2, color = batch)) + 
  geom_point(size = 1.5) + labs(colour = "Batch") +
  theme_minimal()
```


## Feature selection with Limma

```{r}
set.seed(3888)

group_factor = factor(group$Outcome)
design_matrix <- model.matrix(~ 0 + group_factor)
colnames(design_matrix) <- levels(group_factor)
    
contrast_matrix <- makeContrasts(
  Stable_vs_ABMR = Stable - ABMR,
  Stable_vs_TCMR = Stable - TCMR,
  Stable_vs_Rejection = Stable - (1 / 2 * (ABMR + TCMR)),
  levels = colnames(design_matrix)
)
    
fit <- lmFit(combat_eMat, design = design_matrix)
fit <- contrasts.fit(fit, contrasts = contrast_matrix)
fit <- eBayes(fit)
tT <- topTable(fit, n = Inf) %>% arrange('adj.P.Val')

top_100_eMat = combat_eMat[rownames(tT)[1:100],]
```


## MA-plot

```{r}
ggplot(tT, aes(x = AveExpr, y = Stable_vs_ABMR))+
  geom_point(aes(colour = -log10(P.Value)), alpha = 1/3, size = 1) +
  scale_colour_gradient(low = "blue", high = "red")+
  ylab("log2 fold change") + xlab("Average expression") + ggtitle("MA-plot for Stable vs ABMR")

ggplot(tT, aes(x = AveExpr, y = Stable_vs_TCMR))+
  geom_point(aes(colour = -log10(P.Value)), alpha = 1/3, size = 1) +
  scale_colour_gradient(low = "blue", high = "red")+
  ylab("log2 fold change") + xlab("Average expression") + ggtitle("MA-plot for Stable vs TCMR")

ggplot(tT, aes(x = AveExpr, y = Stable_vs_Rejection))+
  geom_point(aes(colour = -log10(P.Value)), alpha = 1/3, size = 1) +
  scale_colour_gradient(low = "blue", high = "red")+
  ylab("log2 fold change") + xlab("Average expression") + ggtitle("MA-plot for Stable vs Rejection")
```


## Volcano plot

```{r}
ggplot(tT, aes(Stable_vs_ABMR, -log10(P.Value))) +
  geom_point(aes(colour = -log10(P.Value)), alpha = 1/3, size = 1) +
  scale_colour_gradient(low = "blue", high = "red") +
  xlab("log2 fold change") + ylab("-log10 p-value") + ggtitle("Volcano plot for Stable vs ABMR")

ggplot(tT, aes(Stable_vs_TCMR, -log10(P.Value))) +
  geom_point(aes(colour = -log10(P.Value)), alpha = 1/3, size = 1) +
  scale_colour_gradient(low = "blue", high = "red") +
  xlab("log2 fold change") + ylab("-log10 p-value") + ggtitle("Volcano plot for Stable vs TCMR")

ggplot(tT, aes(Stable_vs_Rejection, -log10(P.Value))) +
  geom_point(aes(colour = -log10(P.Value)), alpha = 1/3, size = 1) +
  scale_colour_gradient(low = "blue", high = "red") +
  xlab("log2 fold change") + ylab("-log10 p-value") + ggtitle("Volcano plot for Stable vs Rejection")
```


## Heatmap for the top 100 differentially expressed genes selected using Limma

```{r, message=F}
top100_eMat <- combat_eMat[rownames(tT),]
breaks <- seq(min(top100_eMat), max(top100_eMat), length.out = 100)
color_palette <- colorRampPalette(c('dodgerblue4', 'white', 'yellow'))(99)

sample_annotation <- data.frame(Group = factor(group$Outcome))
rownames(sample_annotation) <- colnames(top100_eMat)

annotation_colors <- c(ABMR = 'firebrick4', TCMR = 'darkseagreen3', Stable = 'white')

annotation_col_colors <- list(Group = annotation_colors)

pheatmap(top100_eMat,
         color = color_palette,
         breaks = breaks,
         annotation_col = sample_annotation,
         annotation_colors = annotation_col_colors,
         fontsize = 3,
         fontsize_row = 3,
         fontsize_col = 3,
         clustering_method = "complete",
         scale = "row",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         legend = TRUE)
```


## Model evaluation with cross-validation

```{r}
set.seed(3888)

X = t(combined_expression)
y = factor(group$Outcome)

cvK = 5   
n_sim = 20

cv_accuracy10_rf = cv_accuracy50_rf = cv_accuracy100_rf = numeric(n_sim)
cv_accuracy10_knn = cv_accuracy50_knn = cv_accuracy100_knn = numeric(n_sim)
cv_accuracy10_svm = cv_accuracy50_svm = cv_accuracy100_svm = numeric(n_sim)

for (i in 1:n_sim) {
  cvSets = cvFolds(nrow(X), cvK)
  
  cv_accuracy_folds10_rf = cv_accuracy_folds50_rf = cv_accuracy_folds100_rf = numeric(cvK)
  cv_accuracy_folds10_knn = cv_accuracy_folds50_knn = cv_accuracy_folds100_knn = numeric(cvK)
  cv_accuracy_folds10_svm = cv_accuracy_folds50_svm = cv_accuracy_folds100_svm = numeric(cvK)
  
  for (j in 1:cvK) {
    test_id = cvSets$subsets[cvSets$which == j]
    X_train = X[-test_id,]
    X_test = X[test_id,]
    y_train = y[-test_id]
    y_test = y[test_id]
    
    # Applying batch correction to the training data
    batch_train <- apply(sapply(batch_labels, FUN = `%in%`, x = rownames(X_train)), MARGIN = 1, FUN = which)
    mod <- model.matrix(~1, data = data.frame(ones = rep(1, length(batch_train))))
    X_train <- t(sva::ComBat(dat = t(X_train), batch = batch_train, mod = mod, par.prior = TRUE, prior.plots = FALSE))
    
    # Feature selection with Limma
    design_matrix <- model.matrix(~ 0 + y_train)
    colnames(design_matrix) <- levels(y_train)
    
    contrast_matrix <- makeContrasts(
      Stable_vs_ABMR = Stable - ABMR,
      Stable_vs_TCMR = Stable - TCMR,
      Stable_vs_Rejection = Stable - (1 / 2 * (ABMR + TCMR)),
      levels = colnames(design_matrix)
    )
    
    fit <- lmFit(t(X_train), design = design_matrix)
    fit <- contrasts.fit(fit, contrasts = contrast_matrix)
    fit <- eBayes(fit)
    limma_top <- topTable(fit, n = 100) %>% arrange('adj.P.Val')
    
    DE_genes10 <- rownames(limma_top)[1:10]
    DE_genes50 <- rownames(limma_top)[1:50]
    DE_genes100 <- rownames(limma_top)[1:100]
    
    X_train10 = X_train[,DE_genes10]
    X_test10 = X_test[,DE_genes10]
    
    X_train50 = X_train[,DE_genes50]
    X_test50 = X_test[,DE_genes50]
    
    X_train100 = X_train[,DE_genes100]
    X_test100 = X_test[,DE_genes100]
    
    # Random Forest
    rf_fit10 = randomForest(x = X_train10, y = y_train)
    predictions_rf10 = predict(rf_fit10, X_test10)
    cv_accuracy_folds10_rf[j] = mean(y_test == predictions_rf10)
    
    rf_fit50 = randomForest(x = X_train50, y = y_train)
    predictions_rf50 = predict(rf_fit50, X_test50)
    cv_accuracy_folds50_rf[j] = mean(y_test == predictions_rf50)
    
    rf_fit100 = randomForest(x = X_train100, y = y_train)
    predictions_rf100 = predict(rf_fit100, X_test100)
    cv_accuracy_folds100_rf[j] = mean(y_test == predictions_rf100)
    
    # KNN (k = 5)
    knn_fit10 = class::knn(train = X_train10, test = X_test10, cl = y_train, k = 5)
    cv_accuracy_folds10_knn[j] = mean(y_test == knn_fit10)
    
    knn_fit50 = class::knn(train = X_train50, test = X_test50, cl = y_train, k = 5)
    cv_accuracy_folds50_knn[j] = mean(y_test == knn_fit50)
    
    knn_fit100 = class::knn(train = X_train100, test = X_test100, cl = y_train, k = 5)
    cv_accuracy_folds100_knn[j] = mean(y_test == knn_fit100)
    
    # SVM
    svm_fit10 = svm(X_train10, y_train)
    predictions_svm10 = predict(svm_fit10, X_test10)
    cv_accuracy_folds10_svm[j] = mean(y_test == predictions_svm10)
    
    svm_fit50 = svm(X_train50, y_train)
    predictions_svm50 = predict(svm_fit50, X_test50)
    cv_accuracy_folds50_svm[j] = mean(y_test == predictions_svm50)
    
    svm_fit100 = svm(X_train100, y_train)
    predictions_svm100 = predict(svm_fit100, X_test100)
    cv_accuracy_folds100_svm[j] = mean(y_test == predictions_svm100)
  }
  
  cv_accuracy10_rf[i] = mean(cv_accuracy_folds10_rf)
  cv_accuracy50_rf[i] = mean(cv_accuracy_folds50_rf)
  cv_accuracy100_rf[i] = mean(cv_accuracy_folds100_rf)
  
  cv_accuracy10_knn[i] = mean(cv_accuracy_folds10_knn)
  cv_accuracy50_knn[i] = mean(cv_accuracy_folds50_knn)
  cv_accuracy100_knn[i] = mean(cv_accuracy_folds100_knn)
  
  cv_accuracy10_svm[i] = mean(cv_accuracy_folds10_svm)
  cv_accuracy50_svm[i] = mean(cv_accuracy_folds50_svm)
  cv_accuracy100_svm[i] = mean(cv_accuracy_folds100_svm)
}

boxplot(cv_accuracy10_rf, cv_accuracy50_rf, cv_accuracy100_rf,
        cv_accuracy10_knn, cv_accuracy50_knn, cv_accuracy100_knn,
        cv_accuracy10_svm, cv_accuracy50_svm, cv_accuracy100_svm,
        names = c("RF - 10", "RF - 50", "RF - 100", "KNN - 10", "KNN - 50", "KNN - 100", "SVM - 10", "SVM - 50", "SVM - 100"),
        cex.axis = 0.65)
```


## Final model

```{r}
set.seed(3888)

svm_fit_final = svm(t(top_100_eMat), factor(group$Outcome))
svm_fit_final$fitted
```

