---
title: "DATA3888 Interdisciplinary Project"
author: 'Kidney20'
output:
  html_document:
    fig_caption: yes
    self_contained: yes
    theme: flatly
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
---
# Loading required packages

```{r setup, include=FALSE}
library(caret)
library(cvTools)
library(dplyr)
library(e1071)
library(GEOquery)
library(ggplot2)
library(glmnet)
library(limma)
library(pheatmap)
library(randomForest)
library(stringr)
library(sva)
library(tidyverse)

rm(list = ls())
```


# Loading datasets and cleaning

## 1. GSE36059

```{r}
gse1 <- 'GSE36059'
eSet1 <- getGEO('GSE36059')
eSet1 <- eSet1$GSE36059_series_matrix.txt.gz
exp1 <- exprs(eSet1)
phe1 <- pData(eSet1)
feature1 <- fData(eSet1)
```

### Cleaning feature data

```{r}
f1 <- feature1 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f1$'Gene Symbol' <- sapply(strsplit(f1$'Gene Symbol', ' /// '), function(x) x[1])
```

### Substituting gene probe ID with gene symbol

```{r}
gene_symbols1 <- f1[, c(1, 11)]
expr1 <- as.data.frame(exp1)
expr1[, 'ID'] <- rownames(expr1)
expr1 <- merge(x = expr1, y = gene_symbols1, by.x = 'ID')
```

### Adding index

```{r}
rownames(phe1) <- paste(rownames(phe1), '1', sep='_')
colnames(expr1) <- paste(colnames(expr1), '1', sep='_')
```

### Removing duplicated gene symbols - keeping the mean gene expression

```{r}
expre1 <- limma::avereps(expr1[, -c(1, ncol(expr1))], ID = expr1$`Gene Symbol`)
expression1 <- as.data.frame(expre1)
dim(expression1)
```

### Filtering of outcome groups

```{r}
phe1$Outcome <- ifelse(str_detect(phe1$title, 'non-rejecting'), 'Stable',
                       ifelse(str_detect(phe1$title, 'ABMR'), 'ABMR',
                              ifelse(str_detect(phe1$title, 'TCMR'), 'TCMR', 'Other')))
pheno1 <- phe1 %>% filter(Outcome != 'Other')
table(pheno1$Outcome)

expression1 <- as.data.frame(expression1)
expression1 <- expression1[, colnames(expression1) %in% rownames(pheno1)]
expression1 <- as.matrix(expression1)
```


## 2. GSE129166

```{r}
gse2 <- 'GSE129166'
eSet2 <- getGEO('GSE129166')
eSet2 <- eSet2$GSE129166_series_matrix.txt.gz
exp2 <- exprs(eSet2)
phe2 <- pData(eSet2)
feature2 <- fData(eSet2)
```

### Filtering to biopsy data only

```{r}
phe2 <- phe2 %>% filter(str_detect(title, 'BIOPSY'))
exp2 <- exp2[, rownames(phe2)]
```

### Cleaning feature data

```{r}
f2 <- feature2 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f2$'Gene Symbol' <- sapply(strsplit(f2$'Gene Symbol', ' /// '), function(x) x[1])
```

### Substituting gene probe ID with gene symbol

```{r}
gene_symbols2 <- f2[, c(1, 11)]
expr2 <- as.data.frame(exp2)
expr2[, 'ID'] <- rownames(expr2)
expr2 <- merge(x = expr2, y = gene_symbols2, by.x = 'ID')
```

### Adding index

```{r}
rownames(phe2) <- paste(rownames(phe2), '2', sep='_')
colnames(expr2) <- paste(colnames(expr2), '2', sep='_')
```

### Removing duplicated gene symbols - keeping the mean gene expression

```{r}
expre2 <- limma::avereps(expr2[, -c(1, ncol(expr2))], ID = expr2$`Gene Symbol`)
expression2 <- as.data.frame(expre2)
dim(expression2)
```

### Filtering of outcome groups

```{r}
pheno2 <- phe2 %>% mutate(
  ABMR_Status = ifelse(`abmr (no:ch1` == '0_Yes:1): 1', 'ABMR', 'ABMR Stable'),
  TCMR_Status = ifelse(`tcmr (no:ch1` == '0_borderline:1_TCMR:2): 1', 'TCMR', 'TCMR Stable')
  ) %>%
    filter(!(ABMR_Status == 'ABMR' & TCMR_Status == 'TCMR')) %>%
  mutate(
    Outcome = case_when(
      ABMR_Status == 'ABMR Stable' & TCMR_Status == 'TCMR Stable' ~ 'Stable',
      ABMR_Status == 'ABMR Stable' ~ 'TCMR',
      TCMR_Status == 'TCMR Stable' ~ 'ABMR',
      TRUE ~ NA_character_
    )
  )
table(pheno2$Outcome)

expression2 <- as.data.frame(expression2)
expression2 <- expression2[, colnames(expression2) %in% rownames(pheno2)]
expression2 <- as.matrix(expression2)
```


## 3. GSE192444

```{r}
gse3 <- 'GSE192444'
eSet3 <- getGEO('GSE192444')
eSet3 <- eSet3$GSE192444_series_matrix.txt.gz
exp3 <- exprs(eSet3)
phe3 <- pData(eSet3)
feature3 <- fData(eSet3)
```

### Cleaning feature data

```{r}
f3 <- feature3 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f3$'Gene Symbol' <- sapply(strsplit(f3$'Gene Symbol', ' /// '), function(x) x[1])
```

### Substituting gene probe ID with gene symbol

```{r}
gene_symbols3 <- f3[, c(1, 15)]
expr3 <- as.data.frame(exp3)
expr3[, 'ID'] <- rownames(expr3)
expr3 <- merge(x = expr3, y = gene_symbols3, by.x = 'ID')
```

### Adding index

```{r}
rownames(phe3) <- paste(rownames(phe3), '3', sep='_')
colnames(expr3) <- paste(colnames(expr3), '3', sep='_')
```

### Removing duplicated gene symbols - keeping the mean gene expression

```{r}
expre3 <- limma::avereps(expr3[, -c(1, ncol(expr3))], ID = expr3$`Gene Symbol`)
expression3 <- as.data.frame(expre3)
dim(expression3)
```

### Filtering of outcome groups

```{r}
phe3$Outcome <- ifelse(str_detect(phe3$`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1`, 'ABMR'), 'ABMR',
                       ifelse(str_detect(phe3$`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1`, 'TCMR'), 'TCMR',
                              ifelse(str_detect(phe3$`mmdx (abmr, tcmr, mixed, nr, pabmr, ptcmr):ch1`, 'NR'), 'Stable', 'Other')))
pheno3 <- phe3 %>% filter(Outcome != 'Other')
table(pheno3$Outcome)

expression3 <- as.data.frame(expression3)
expression3 <- expression3[, colnames(expression3) %in% rownames(pheno3)]
expression3 <- as.matrix(expression3)
```


## 4. GSE124203

```{r}
gse4 <- 'GSE124203'
eSet4 <- getGEO('GSE124203')
eSet4 <- eSet4$GSE124203_series_matrix.txt.gz
exp4 <- exprs(eSet4)
phe4 <- pData(eSet4)
feature4 <- fData(eSet4)
```

### Cleaning feature data

```{r}
f4 <- feature4 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f4$'Gene Symbol' <- sapply(strsplit(f4$'Gene Symbol', ' /// '), function(x) x[1])
```

### Substituting gene probe ID with gene symbol

```{r}
gene_symbols4 <- f4[, c(1, 15)]
expr4 <- as.data.frame(exp4)
expr4[, 'ID'] <- rownames(expr4)
expr4 <- merge(x = expr4, y = gene_symbols4, by.x = 'ID')
```

### Adding index

```{r}
rownames(phe4) <- paste(rownames(phe4), '4', sep='_')
colnames(expr4) <- paste(colnames(expr4), '4', sep='_')
```

### Removing duplicated gene symbols - keeping the mean gene expression

```{r}
expre4 <- limma::avereps(expr4[, -c(1, ncol(expr4))], ID = expr4$`Gene Symbol`)
expression4 <- as.data.frame(expre4)
dim(expression4)
```

### Filtering of outcome groups

```{r}
phe4$Outcome <- ifelse(str_detect(phe4$`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1`, 'ABMR'), 'ABMR',
                       ifelse(str_detect(phe4$`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1`, 'TCMR'), 'TCMR',
                              ifelse(str_detect(phe4$`mmdx (abmr, pabmr, mixed, tcmr, ptcmr, nr, na):ch1`, 'NR'), 'Stable', 'Other')))
pheno4 <- phe4 %>% filter(Outcome != 'Other')
table(pheno4$Outcome)

expression4 <- as.data.frame(expression4)
expression4 <- expression4[, colnames(expression4) %in% rownames(pheno4)]
expression4 <- as.matrix(expression4)
```


## 5. GSE98320

```{r}
gse5 <- 'GSE98320'
eSet5 <- getGEO('GSE98320')
eSet5 <- eSet5$GSE98320_series_matrix.txt.gz
exp5 <- exprs(eSet5)
phe5 <- pData(eSet5)
feature5 <- fData(eSet5)
```

### Cleaning feature data

```{r}
f5 <- feature5 %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f5$'Gene Symbol' <- sapply(strsplit(f5$'Gene Symbol', ' /// '), function(x) x[1])
```

### Substituting gene probe ID with gene symbol

```{r}
gene_symbols5 <- f5[, c(1, 17)]
expr5 <- as.data.frame(exp5)
expr5[, 'ID'] <- rownames(expr5)
expr5 <- merge(x = expr5, y = gene_symbols5, by.x = 'ID')
```

### Adding index

```{r}
rownames(phe5) <- paste(rownames(phe5), '5', sep='_')
colnames(expr5) <- paste(colnames(expr5), '5', sep='_')
```

### Removing duplicated gene symbols - keeping the mean gene expression

```{r}
expre5 <- limma::avereps(expr5[,-c(1, ncol(expr5))], ID = expr5$`Gene Symbol`)
expression5 <- as.data.frame(expre5)
dim(expression5)
```

### Filtering of outcome groups

```{r}
pheno5 <- phe5 %>% 
  filter(`mmdx:ch1` != '-', `mmdx:ch1` != 'Mixed')%>% 
  mutate(Outcome = case_when(
  `mmdx:ch1` == 'ABMR' ~ 'ABMR',
  `mmdx:ch1` == 'TCMR' ~ 'TCMR',
  `mmdx:ch1` == 'NR' ~ 'Stable',
  TRUE ~ NA_character_
))
table(pheno5$Outcome)

expression5 <- as.data.frame(expression5)
expression5 <- expression5[, colnames(expression5) %in% rownames(pheno5)]
expression5 <- as.matrix(expression5)
```


# Evaluation of datasets, prediction with models

## Selecting the top 75 percentile gene expressions

```{r}
top_75_percentile <- function(eMat) {
  eMat <- as.data.frame(eMat)
  threshold <- quantile(rowSums(eMat), 0.25)
  filtered_eMat <- eMat[rowSums(eMat) > threshold,]
  return (filtered_eMat)
}

expression1 <- top_75_percentile(expression1)
expression2 <- top_75_percentile(expression2)
expression3 <- top_75_percentile(expression3)
expression4 <- top_75_percentile(expression4)
expression5 <- top_75_percentile(expression5)

pheno1 <- pheno1[colnames(expression1),]
pheno2 <- pheno2[colnames(expression2),]
pheno3 <- pheno3[colnames(expression3),]
pheno4 <- pheno4[colnames(expression4),]
pheno5 <- pheno5[colnames(expression5),]
```


## Combining the different datasets

```{r,message=F}
common_genes <- Reduce(intersect, list(rownames(expression1), rownames(expression2), 
                                       rownames(expression3), rownames(expression4), 
                                       rownames(expression5)))

expression1_common <- expression1[common_genes,]
expression2_common <- expression2[common_genes,]
expression3_common <- expression3[common_genes,]
expression4_common <- expression4[common_genes,]
expression5_common <- expression5[common_genes,]

combined_expression <- cbind(expression1_common, expression2_common, 
                             expression3_common, expression4_common, 
                             expression5_common)
dim(combined_expression)
```


## Adding batch labels

```{r}
# Create a list where each element is a vector of column names for each dataset
batch_labels <- list(
  expression1 = colnames(expression1_common),
  expression2 = colnames(expression2_common),
  expression3 = colnames(expression3_common),
  expression4 = colnames(expression4_common),
  expression5 = colnames(expression5_common)
)
names(batch_labels) <- c("1", "2", "3", "4", "5")
```


## Combined outcomes

```{r}
pheno1 <- pheno1[colnames(expression1_common),]
pheno2 <- pheno2[colnames(expression2_common),]
pheno3 <- pheno3[colnames(expression3_common),]
pheno4 <- pheno4[colnames(expression4_common),]
pheno5 <- pheno5[colnames(expression5_common),]

group1 <- data.frame(Outcome = pheno1$Outcome, row.names = rownames(pheno1))
group2 <- data.frame(Outcome = pheno2$Outcome, row.names = rownames(pheno2))
group3 <- data.frame(Outcome = pheno3$Outcome, row.names = rownames(pheno3))
group4 <- data.frame(Outcome = pheno4$Outcome, row.names = rownames(pheno4))
group5 <- data.frame(Outcome = pheno5$Outcome, row.names = rownames(pheno5))

group <- dplyr::bind_rows(group1, group2, group3, group4, group5)
table(group)
```


## Balancing of outcome groups and normalising expression intensities

```{r}
set.seed(3888)

# Aim: ABMR = TCMR = Stable = 227
# Picking random samples
ABMR_samples <- sample(colnames(combined_expression)[group == 'ABMR'], 227)
TCMR_samples <- sample(colnames(combined_expression)[group == 'TCMR'], 227)
Stable_samples <- sample(colnames(combined_expression)[group == 'Stable'], 227)
selected_samples <- c(ABMR_samples, TCMR_samples, Stable_samples)

combined_expression <- combined_expression[, selected_samples] 
combined_expression <- normalizeBetweenArrays(combined_expression)

filter_batches <- function(batches, samples) {
  lapply(batches, function(batch) batch[batch %in% samples])
}

batch_labels <- filter_batches(batch_labels, selected_samples)

group <- group[selected_samples, , drop = FALSE]
batch_info <- apply(sapply(batch_labels, FUN = `%in%`, x = rownames(group)), MARGIN = 1, FUN = which)

table(group)
dim(combined_expression) # n = 227 * 3 = 681
# print(lapply(batch_labels, length))
```


## Exploring data with PCA - before batch correction

```{r}
set.seed(3888)

gse_pca <- prcomp(t(combined_expression))
df_toplot <- data.frame(Outcome = group$Outcome,
                        pc1 = gse_pca$x[,1], pc2 = gse_pca$x[,2],
                        batch = batch_info)

ggplot(df_toplot, aes(x = pc1, y = pc2, color = Outcome)) + 
  geom_point(size = 1.5) + labs(colour = "Outcome") +
  theme_minimal()

ggplot(df_toplot, aes(x = pc1, y = pc2, color = factor(batch))) + 
  geom_point(size = 1.5) + labs(colour = "Batch") +
  theme_minimal()
```


## ComBat batch correction

```{r, echo=FALSE}
MOD <- model.matrix(~factor(group$Outcome))
combat_eMat <- sva::ComBat(dat = combined_expression, batch = batch_info, mod = MOD, par.prior = TRUE, prior.plots = FALSE) 
dim(combat_eMat)
```


## Exploring data with PCA - after batch correction

```{r}
set.seed(3888)

gse_pca <- prcomp(t(combat_eMat))
df_toplot <- data.frame(Outcome = group$Outcome,
                        pc1 = gse_pca$x[,1], pc2 = gse_pca$x[,2],
                        batch = as.factor(batch_info))

ggplot(df_toplot, aes(x = pc1, y = pc2, color = Outcome)) + 
  geom_point(size = 1.5) + labs(colour = "Outcome") +
  theme_minimal()

ggplot(df_toplot, aes(x = pc1, y = pc2, color = batch)) + 
  geom_point(size = 1.5) + labs(colour = "Batch") +
  theme_minimal()
```


## Feature selection with Limma

### Top genes
```{r}
set.seed(3888)

limma_top_genes <- function(eMat, outcome, n) {
  group_factor = factor(outcome)
  design_matrix <- model.matrix(~ 0 + group_factor)
  colnames(design_matrix) <- levels(group_factor)
      
  contrast_matrix <- makeContrasts(
    ABMR_vs_Stable_and_TCMR = ABMR - (1 / 2 * (Stable + TCMR)),
    TCMR_vs_Stable_and_ABMR = TCMR - (1 / 2 * (Stable + ABMR)),
    Stable_vs_Rejection = Stable - (1 / 2 * (ABMR + TCMR)),
    levels = colnames(design_matrix)
  )
      
  fit <- lmFit(eMat, design = design_matrix)
  fit <- contrasts.fit(fit, contrasts = contrast_matrix)
  fit <- eBayes(fit)
  
  tT_ABMR <- topTable(fit, n = n, coef = 1)
  tT_TCMR <- topTable(fit, n = n, coef = 2)
  tT_Stable <- topTable(fit, n = n, coef = 3)

  top_genes <- unique(c(rownames(tT_ABMR), rownames(tT_TCMR), rownames(tT_Stable)))
  
  return (top_genes)
}

top_genes = limma_top_genes(combat_eMat, group$Outcome, 17)#48 genes
# top_genes_eMat = combat_eMat[top_genes,]
```

### Limma for plots
```{r}
group_factor_plot = factor(group$Outcome)
  design_matrix_plot <- model.matrix(~ 0 + group_factor_plot)
  colnames(design_matrix_plot) <- levels(group_factor_plot)
      
  contrast_matrix_plot <- makeContrasts(
    ABMR_vs_Stable_and_TCMR = ABMR - (1 / 2 * (Stable + TCMR)),
    TCMR_vs_Stable_and_ABMR = TCMR - (1 / 2 * (Stable + ABMR)),
    Stable_vs_Rejection = Stable - (1 / 2 * (ABMR + TCMR)),
    levels = colnames(design_matrix_plot)
  )
      
fit_plot <- lmFit(combat_eMat, design = design_matrix_plot)
fit_plot <- contrasts.fit(fit_plot, contrasts = contrast_matrix_plot)
fit_plot <- eBayes(fit_plot)
tT_plot <- topTable(fit_plot, n = Inf)
```

## MA-plot

```{r}
ggplot(tT_plot, aes(x = AveExpr, y = ABMR_vs_Stable_and_TCMR))+
  geom_point(aes(colour = -log10(P.Value)), alpha = 1/3, size = 1) +
  scale_colour_gradient(low = "blue", high = "red")+
  ylab("log2 fold change") + xlab("Average expression") + ggtitle("MA-plot for SABMR_vs_Stable_and_TCMR")

ggplot(tT_plot, aes(x = AveExpr, y = TCMR_vs_Stable_and_ABMR))+
  geom_point(aes(colour = -log10(P.Value)), alpha = 1/3, size = 1) +
  scale_colour_gradient(low = "blue", high = "red")+
  ylab("log2 fold change") + xlab("Average expression") + ggtitle("MA-plot for TCMR_vs_Stable_and_ABMR")

ggplot(tT_plot, aes(x = AveExpr, y = Stable_vs_Rejection))+
  geom_point(aes(colour = -log10(P.Value)), alpha = 1/3, size = 1) +
  scale_colour_gradient(low = "blue", high = "red")+
  ylab("log2 fold change") + xlab("Average expression") + ggtitle("MA-plot for Stable_vs_Rejection")
```


## Volcano plot

```{r}
ggplot(tT_plot, aes(ABMR_vs_Stable_and_TCMR, -log10(P.Value))) +
  geom_point(aes(colour = -log10(P.Value)), alpha = 1/3, size = 1) +
  scale_colour_gradient(low = "blue", high = "red") +
  xlab("log2 fold change") + ylab("-log10 p-value") + ggtitle("Volcano plot for ABMR_vs_Stable_and_TCMR")

ggplot(tT_plot, aes(TCMR_vs_Stable_and_ABMR, -log10(P.Value))) +
  geom_point(aes(colour = -log10(P.Value)), alpha = 1/3, size = 1) +
  scale_colour_gradient(low = "blue", high = "red") +
  xlab("log2 fold change") + ylab("-log10 p-value") + ggtitle("Volcano plot for TCMR_vs_Stable_and_ABMR")

ggplot(tT_plot, aes(Stable_vs_Rejection, -log10(P.Value))) +
  geom_point(aes(colour = -log10(P.Value)), alpha = 1/3, size = 1) +
  scale_colour_gradient(low = "blue", high = "red") +
  xlab("log2 fold change") + ylab("-log10 p-value") + ggtitle("Volcano plot for Stable vs Rejection")
```


## Heatmap for the DE genes selected using Limma & Lasso

### Lasso
```{r}
set.seed(3888)
genes <- combat_eMat[rownames(tT_plot),]
group_numeric <- as.numeric(factor(group$Outcome))
x<- t(genes)
cv_lasso <- cv.glmnet(x, group_numeric, alpha = 1)
opt_lambda <- cv_lasso$lambda.1se
final_lasso <- glmnet(x, group_numeric, alpha = 1, lambda = opt_lambda)
lasso_coef <- coef(final_lasso, s = opt_lambda)
selected_genes_heatmap <- rownames(genes)[which(lasso_coef[-1,] != 0)]
selected_genes_heatmap# 45 genes
```
### Heatmap for top 45 genes
```{r, message=F}
normalised_top49 <- scale(combat_eMat[selected_genes_heatmap,])
sample_annotation <- data.frame(Group = factor(group$Outcome))
rownames(sample_annotation) <- colnames(normalised_top49)
pheatmap(normalised_top49,
         annotation_col = sample_annotation,
         fontsize = 3,
         fontsize_row = 3,
         fontsize_col = 5,
         clustering_method = "complete",
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         scale = "row",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         legend = TRUE,
         main = "Heatmap of Selected Genes")
```


## Model evaluation with cross-validation

```{r}
set.seed(3888)

X = t(combined_expression)
y = factor(group$Outcome)

cvK = 5   
n_sim = 20

cv_accuracy30_rf = cv_accuracy50_rf = cv_accuracy100_rf = numeric(n_sim)
cv_accuracy30_knn2 = cv_accuracy50_knn2 = cv_accuracy100_knn2 = numeric(n_sim)
cv_accuracy30_knn5 = cv_accuracy50_knn5 = cv_accuracy100_knn5 = numeric(n_sim)
cv_accuracy30_svm = cv_accuracy50_svm = cv_accuracy100_svm = numeric(n_sim)

for (i in 1:n_sim) {
  cvSets = cvFolds(nrow(X), cvK)
  
  cv_accuracy_folds30_rf = cv_accuracy_folds50_rf = cv_accuracy_folds100_rf = numeric(cvK)
  cv_accuracy_folds30_knn2 = cv_accuracy_folds50_knn2 = cv_accuracy_folds100_knn2 = numeric(cvK)
  cv_accuracy_folds30_knn5 = cv_accuracy_folds50_knn5 = cv_accuracy_folds100_knn5 = numeric(cvK)
  cv_accuracy_folds30_svm = cv_accuracy_folds50_svm = cv_accuracy_folds100_svm = numeric(cvK)
  
  
  for (j in 1:cvK) {
    test_id = cvSets$subsets[cvSets$which == j]
    X_train = X[-test_id,]
    X_test = X[test_id,]
    y_train = y[-test_id]
    y_test = y[test_id]
    
    # Applying batch correction to the training data
    batch_train <- apply(sapply(batch_labels, FUN = `%in%`, x = rownames(X_train)), MARGIN = 1, FUN = which)
    mod <- model.matrix(~factor(group[rownames(X_train),]))
    X_train <- t(sva::ComBat(dat = t(X_train), batch = batch_train, mod = mod, par.prior = TRUE, prior.plots = FALSE))
    
    # Feature selection with Limma
    DE_genes30 <- limma_top_genes(t(X_train), y_train, 10)
    DE_genes50 <- limma_top_genes(t(X_train), y_train, 17)
    DE_genes100 <- limma_top_genes(t(X_train), y_train, 33)
    
    X_train30 = X_train[,DE_genes30]
    X_test30 = X_test[,DE_genes30]
    
    X_train50 = X_train[,DE_genes50]
    X_test50 = X_test[,DE_genes50]
    
    X_train100 = X_train[,DE_genes100]
    X_test100 = X_test[,DE_genes100]
    
    
    # Random Forest
    rf_fit30 = randomForest(x = X_train30, y = y_train)
    predictions_rf30 = predict(rf_fit30, X_test30)
    cv_accuracy_folds30_rf[j] = mean(y_test == predictions_rf30)
    
    rf_fit50 = randomForest(x = X_train50, y = y_train)
    predictions_rf50 = predict(rf_fit50, X_test50)
    cv_accuracy_folds50_rf[j] = mean(y_test == predictions_rf50)
    
    rf_fit100 = randomForest(x = X_train100, y = y_train)
    predictions_rf100 = predict(rf_fit100, X_test100)
    cv_accuracy_folds100_rf[j] = mean(y_test == predictions_rf100)
    
    # SVM
    svm_fit30 = svm(X_train30, y_train)
    predictions_svm30 = predict(svm_fit30, X_test30)
    cv_accuracy_folds30_svm[j] = mean(y_test == predictions_svm30)
    
    svm_fit50 = svm(X_train50, y_train)
    predictions_svm50 = predict(svm_fit50, X_test50)
    cv_accuracy_folds50_svm[j] = mean(y_test == predictions_svm50)
    
    svm_fit100 = svm(X_train100, y_train)
    predictions_svm100 = predict(svm_fit100, X_test100)
    cv_accuracy_folds100_svm[j] = mean(y_test == predictions_svm100)
    
    # KNN (k = 2)
     knn2_fit30 = class::knn(train = X_train30, test = X_test30, cl = y_train, k = 2)
     cv_accuracy_folds30_knn2[j] = mean(y_test == knn2_fit30)
     
     knn2_fit50 = class::knn(train = X_train50, test = X_test50, cl = y_train, k = 2)
     cv_accuracy_folds50_knn2[j] = mean(y_test == knn2_fit50)
     
     knn2_fit100 = class::knn(train = X_train100, test = X_test100, cl = y_train, k = 2)
     cv_accuracy_folds100_knn2[j] = mean(y_test == knn2_fit100)
    
    # KNN (k = 5)
     knn5_fit30 = class::knn(train = X_train30, test = X_test30, cl = y_train, k = 5)
     cv_accuracy_folds30_knn5[j] = mean(y_test == knn5_fit30)
     
     knn5_fit50 = class::knn(train = X_train50, test = X_test50, cl = y_train, k = 5)
     cv_accuracy_folds50_knn5[j] = mean(y_test == knn5_fit50)
     
     knn5_fit100 = class::knn(train = X_train100, test = X_test100, cl = y_train, k = 5)
     cv_accuracy_folds100_knn5[j] = mean(y_test == knn5_fit100)
  }
  
  cv_accuracy30_rf[i] = mean(cv_accuracy_folds30_rf)
  cv_accuracy50_rf[i] = mean(cv_accuracy_folds50_rf)
  cv_accuracy100_rf[i] = mean(cv_accuracy_folds100_rf)
  
  cv_accuracy30_svm[i] = mean(cv_accuracy_folds30_svm)
  cv_accuracy50_svm[i] = mean(cv_accuracy_folds50_svm)
  cv_accuracy100_svm[i] = mean(cv_accuracy_folds100_svm)
  
  cv_accuracy30_knn2[i] = mean(cv_accuracy_folds30_knn2)
  cv_accuracy50_knn2[i] = mean(cv_accuracy_folds50_knn2)
  cv_accuracy100_knn2[i] = mean(cv_accuracy_folds100_knn2)
 
  cv_accuracy30_knn5[i] = mean(cv_accuracy_folds30_knn5)
  cv_accuracy50_knn5[i] = mean(cv_accuracy_folds50_knn5)
  cv_accuracy100_knn5[i] = mean(cv_accuracy_folds100_knn5)
}

 boxplot(cv_accuracy30_rf, cv_accuracy50_rf, cv_accuracy100_rf,
         cv_accuracy30_svm, cv_accuracy50_svm, cv_accuracy100_svm,
         cv_accuracy30_knn2, cv_accuracy50_knn2, cv_accuracy100_knn2,
         cv_accuracy30_knn5, cv_accuracy50_knn5, cv_accuracy100_knn5,
        names = c("RF - 30", "RF - 50", "RF - 100", 'SVM -30','SVM -50','SVM -100',"KNN2 - 30", "KNN2 - 50", "KNN2 - 100", "KNN5 - 30", "KNN5 - 50", "KNN5 - 100"),las=2,
         cex.axis = 0.65)

```
#### Evaluation of cv
```{r}
median_rf_50 <- median(cv_accuracy50_rf)
median_rf_100 <- median(cv_accuracy100_rf)

median_rf_50
median_rf_100
```

# Final model (rf50)
## Preparation 

### Loading tetsing data set
```{r}
testing_data <-'GSE48581'
eSet <- getGEO(testing_data)
eSet <- eSet$GSE48581_series_matrix.txt.gz
exp <- exprs(eSet)
pheotype <- pData(eSet)
feature <- fData(eSet)
```
### Cleaning feature
```{r}
f <- feature %>% filter(`Gene Symbol` != '', !is.na(`Gene Symbol`))
f$'Gene Symbol' <- sapply(strsplit(f$'Gene Symbol', ' /// '), function(x) x[1])
```

### Substituting gene probe ID with gene symbol

```{r}
gene_symbols <- f[, c(1, 11)]
expr <- as.data.frame(exp)
expr[, 'ID'] <- rownames(expr)
expr <- merge(x = expr, y = gene_symbols, by.x = 'ID')
```

### Removing duplicated gene symbols - keeping the mean gene expression

```{r}
expre <- limma::avereps(expr[, -c(1, ncol(expr))], ID = expr$`Gene Symbol`)
expression <- as.data.frame(expre)
dim(expression)
```

### Filtering of outcome groups

```{r}
pheotype_filtered <- pheotype %>%
  filter(`diagnosis (tcmr,non-tcmr, nephrectomies):ch1` != 'Nephrectomies') %>%
  filter(!(`histologic diagnosis (abmr, tcmr, bdl, mixed, others):ch1` %in% c('NA', 'Borderline', 'Mixed'))) %>%
  mutate(Outcome = case_when(
    `histologic diagnosis (abmr, tcmr, bdl, mixed, others):ch1` == 'ABMR' ~ 'ABMR',
    `histologic diagnosis (abmr, tcmr, bdl, mixed, others):ch1` == 'TCMR' ~ 'TCMR',
    `histologic diagnosis (abmr, tcmr, bdl, mixed, others):ch1` == 'Others' ~ 'Stable'
  ))

table(pheotype_filtered$Outcome)
```

### Alignment
```{r}
expression <- as.data.frame(expression)
expression_1<- expression[, colnames(expression) %in% rownames(pheotype_filtered)]
expression_2 <- as.matrix(expression_1)
```

### Clarification
```{r}
#eMat_testing <- normalizeBetweenArrays(expression_2)
eMat_testing <- expression_2
```

## Final Model

### rf50
```{r}
set.seed(3888)
group_factor = factor(group$Outcome)
design_matrix <- model.matrix(~ 0 + group_factor)
colnames(design_matrix) <- levels(group_factor)
      
contrast_matrix <- makeContrasts(
    ABMR_vs_Stable_and_TCMR = ABMR - (1 / 2 * (Stable + TCMR)),
    TCMR_vs_Stable_and_ABMR = TCMR - (1 / 2 * (Stable + ABMR)),
    Stable_vs_Rejection = Stable - (1 / 2 * (ABMR + TCMR)),
    levels = colnames(design_matrix)
  )

Fit <- lmFit(combat_eMat, design = design_matrix)
Fit <- contrasts.fit(Fit, contrasts = contrast_matrix)
Fit <- eBayes(Fit)
  
tT_ABMR <- topTable(Fit, n = 17, coef = 1)%>% arrange('adj.P.Val')
tT_TCMR <- topTable(Fit, n = 17, coef = 2)%>% arrange('adj.P.Val')
tT_Stable <- topTable(Fit, n = 17, coef = 3)%>% arrange('adj.P.Val')

top_genes <- unique(c(rownames(tT_ABMR), rownames(tT_TCMR), rownames(tT_Stable)))#48
  
top_48_eMat <- combat_eMat[rownames(combat_eMat) %in% top_genes,]
top_48_eMat <- normalizeBetweenArrays(top_48_eMat)
X_test48 <- eMat_testing[rownames(eMat_testing) %in% top_genes, ]
X_test48 <- t(normalizeBetweenArrays(X_test48))
X_train48 <- t(top_48_eMat)
y_train48 <- factor(group[rownames(X_train48),])
y_test48 <-factor(pheotype_filtered[rownames(X_test48),]$Outcome)

rf48 <- randomForest(x=X_train48,y=y_train48)
predictions_final48 <- predict(rf48, newdata=X_test48)

confusionMatrix48 <- table(Predicted = predictions_final48, Actual = y_test48)
accuracy48 <- sum(diag(confusionMatrix48)) / sum(confusionMatrix48)
confusionMatrix48
accuracy48
```
```{r}
saveRDS(rf48, file = "model.rds")
```

```{r}
X_test48
```





